---
output: github_document
---

```{r, echo = FALSE}
set.seed(17760701)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# randomizr: Easy to Use Tools for Common Forms of Random Assignment and Sampling

**randomizr** is designed to make conducting field, lab, survey, or online experiments easier by automating the random assignment process. Social and lab scientists conducting experiments need a process to assign individuals or units of observation to treatment or control wings. Common designs include simple random assignment, complete randomization, block randomization, cluster randomization, and blocked cluster randomization. **randomizr** automates all of these processes and assists scientists in doing transparent, replicable science.

# Installing randomizr

Installing the latest production-ready version of **randomizr** in `R` is simple.

```{r,eval=FALSE}
install.packages("randomizr")
```

If you would like to use the current development release of **randomizr** (please be aware that development releases may include bugs or syntax changes), run the following code:

```{r,eval=FALSE}
install.packages("devtools")
devtools::install_github("DeclareDesign/randomizr")
```

# Getting started

**randomizr** has five main random assignment functions, corresponding to the common experimental designs listed above. You can read more about using each of these functions in our [reference library](reference/index.html) or by clicking on the function names: `simple_ra()`, `complete_ra()`, `block_ra()`, `cluster_ra()`, and `block_and_cluster_ra()`.

`complete_ra()` (Complete randomization) is the function that will be most appropriate for a large number of experimental situations: when you want to assign a fixed `m` units out of a population of `N` units to treatment:

```{r}
library(randomizr)
Z <- complete_ra(N = 100, m = 50)
table(Z)
```

A more complicated design that, for example, assigns different numbers of clusters to three different treatments, makes use of `cluster_ra()` (Cluster randomization):

```{r}
# This makes a cluster variable: one unit in cluster "a", two in "b"...
clust_var <- rep(letters, times = 1:26)

Z <- cluster_ra(
  clust_var = clust_var,
  m_each = c(7, 7, 12),
  condition_names = c("control", "placebo", "treatment")
  )
table(Z, clust_var)
```

For more information about all of **randomizr**'s functionality, please see our [online tutorial](articles/randomizr_vignette.html)

Happy randomizing!
