---
output: github_document
title: "randomizr: Tools for random assignment and random sampling"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

> **You are on the `rewrite` branch.** This is the 2.0 rewrite of randomizr, not the CRAN package and not `main`. The released randomizr is unaffected by anything here.
>
> Every interface the released package exports is unchanged, so existing code runs unchanged. It adds two exports, `balanced_ra()` and `balanced_ra_probabilities()`, for assignment with unit-varying probabilities; both are experimental. Internally it adds an Rcpp Knuth-shuffle fast path for two-arm block assignment, declaration caching, and `mget`-based S3 dispatch.
>
> ```r
> remotes::install_github("DeclareDesign/randomizr@rewrite")
> ```
>
> Sibling branches: `DeclareDesign/fabricatr@rewrite`, `DeclareDesign/estimatr@rewrite`, `DeclareDesign/DeclareDesign@rewrite`.

```{r, echo = FALSE}
set.seed(42)
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>",
  fig.path = "README-"  
)
options(digits = 2)
```

[![CRAN status](https://www.r-pkg.org/badges/version/randomizr)](https://cran.r-project.org/package=randomizr)
[![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/grand-total/randomizr?color=green)](https://r-pkg.org/pkg/randomizr)
[![Build status](https://github.com/DeclareDesign/randomizr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/DeclareDesign/randomizr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/DeclareDesign/randomizr/graph/badge.svg)](https://app.codecov.io/gh/DeclareDesign/randomizr)
[![Replications](https://softwarecite.com/badge/randomizr)](https://softwarecite.com/package/randomizr)

**randomizr** generates random assignments for common experimental designs, including simple random assignment, complete random assignment, block random assignment, and cluster random assignment.

### Installing randomizr

Use the following to install the latest CRAN release of **randomizr**:

```{r,eval=FALSE}
install.packages("randomizr")
```

### Getting started with randomizr

**randomizr** has four main random assignment functions, corresponding to the common experimental designs listed above. You can read more about using each of these functions in our [reference library](https://declaredesign.org/r/randomizr/reference/) or by clicking on the function names: `simple_ra()`, `complete_ra()`, `block_ra()`, and `cluster_ra()`.

`complete_ra()`: Under complete random assignment, we assign a fixed `m` units out of a population of `N` units to treatment:

```{r echo=TRUE, results="hide"}
library(randomizr)
Z <- complete_ra(N = 100, m = 50)
table(Z)
```
```{r echo=FALSE}
knitr::kable(t(as.matrix(table(Z))))
```

`cluster_ra()`: Under cluster random assignment, whole clusters of units (like all the students in a classroom or everyone living in the same household) are assigned to treatment conditions together.

```{r echo=TRUE, results="hide"}
# This makes a cluster variable: one unit in cluster "a", two in "b"...
clust_var <- rep(letters[1:15], times = 1:15)

Z <- cluster_ra(
  clusters = clust_var,
  m_each = c(4, 4, 7),
  conditions = c("control", "placebo", "treatment")
  )
table(Z, clust_var)
```
```{r echo=FALSE}
knitr::kable(table(Z, clust_var))
```

For more information about all of **randomizr**'s functionality, please see our [online tutorial](https://declaredesign.org/r/randomizr/articles/randomizr_vignette.html)

Happy randomizing!
