---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

randomizr is designed to make conducting field, lab, survey, or online experiments easier by automating the random assignment process.  

Installation from CRAN is easy:

```{r,eval=FALSE}
install.packages("randomizr")
```

If you'd like to install the most current development release, use the following code:

```{r,eval=FALSE}
install.packages("devtools")
devtools::install_github("acoppock/randomizr")
```

There are five main functions in randomizr: `simple_ra()`, `complete_ra()`, `block_ra()`, `cluster_ra()`, and `block_and_cluster_ra()`, which correspond to common experimental designs. 

`complete_ra()` is the workhorse function that will be most appropriate for a large number of experimental situations: it assigns m of N units to treatment:

```{r}
library(randomizr)
Z <- complete_ra(N = 100, m = 50)
table(Z)
```

A more complicated design that, for example, assigns different numbers of clusters to three different treatments can be accomodated like this:

```{r}
# This makes a cluster variable: one unit in cluster "a", two in "b"...
clust_var <- rep(letters, times = 1:26)

Z <- cluster_ra(clust_var=clust_var, m_each = c(7, 7, 12),
                condition_names=c("control", "placebo", "treatment"))
table(Z, clust_var)
```

Happy randomizing!
