---
title: "Introduction to balanced_ra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to balanced_ra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
set.seed(20260822)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
options(digits = 3)
```

```{r}
library(randomizr)
library(ggplot2)
```

`balanced_ra()` draws random assignment with tight targets: condition counts at the floor or the ceiling of what the probabilities imply, and, with `formula`, at the floor or ceiling of covariate totals. Each unit's probability stays exact. Probabilities may vary from unit to unit and block and cluster structures are respected.  With two arms and `blocks`, those tight counts are both within each block and overall. With three or more arms and `blocks`, tightness is within each block; the overall count is the sum of independent block landings and can move by more than one.

Section 1 gives a blocked design that complete assignment inside each block cannot hit. Section 2 shows randomization against a continuous covariate. Section 3 names the three C++ kernels. Sections 4--6 walk through them: the two-arm pivot, the multi-arm cube, and cube-on-X. Section 7 states what this implementation guarantees. Section 8 gives further examples. Section 9 reports a concise validation. Section 10 lists the caveats.

The function is **experimental**. Use `balanced_ra` or declare it with `declare_ra(..., ra_type = "balanced")`, by supplying `prob_unit_each`, or by supplying `formula`.

# 1. Demonstration 1: Block structures and global targets

We first illustrate with a 6 unit, two-block design, with 3 units to be assigned to treatment with uniform probabilities. This is a simple case that complete random assignment inside each block cannot handle. Two districts of three villages, three villages to treat, equal probabilities, blocked by district: the per-district target is 1.5. Each district should receive one or two treated villages, never zero and never three, and the total should be three on every draw.

```{r}
blocks <- rep(1:2, each = 3)
balanced_ra(blocks = blocks)
```

Repeating the draw, the unit means sit at 0.5, each district contributes 1 or 2, and the total is always 3.

```{r}
reps <- replicate(5000, balanced_ra(blocks = blocks))

# individual assignment probabilities
cbind(target = .5, average = rowMeans(reps))

# block totals
table(colSums(reps[blocks == 1, ]), colSums(reps[blocks == 2, ]))
```

# 2. Demonstration 2: A continuous covariate

Blocks tighten counts inside discrete groups. A continuous covariate is a different target: the treated total of $x$ should sit near what the assignment probabilities imply.

Pass `formula = ~ x`. The model matrix is the balancing matrix $X$ in the cube method of [Deville and Tillé (2004)](https://doi.org/10.1093/biomet/91.4.893). The intercept is the count constraint. The $x$ column asks $\sum_i x_i Z_i$ to stay near $\sum_i x_i p_i$. `~ 0 + x` drops the intercept, and the treated count may wander. `blocks` and `formula` cannot be combined.

Four units, $x = (1, 2, 3, 6)$, two assigned to treatment, $p_i = 1/2$. There are six assignments. The target treated count is 2. The target treated *total* of $x$ is 6 (that is $.5 \sum x = 6$). But notice that no possible assignment hits 6: the six pair-sums are 3, 4, 5, 7, 8, and 9.

`complete_ra(N = 4, m = 2)` puts equal weight on each of those six (the sum of $x$ is 6 in expectation, with large variance). `balanced_ra(formula = ~ x)` does not. Both keep every unit's probability at $1/2$ and treat exactly two units on every draw. They differ in the joint distribution.

```{r}
x <- c(1, 2, 3, 6)
pairs <- combn(4, 2)
assign_id <- function(z) paste(which(z == 1), collapse = ",")

n_draw <- 8000
set.seed(20260822)
Z_complete <- replicate(n_draw, complete_ra(N = 4, m = 2))
Z_balanced <- replicate(n_draw, balanced_ra(formula = ~ x))

share <- function(Z) {
  ids <- apply(pairs, 2, paste, collapse = ",")
  tab <- table(factor(apply(Z, 2, assign_id), levels = ids))
  as.numeric(tab) / ncol(Z)
}

tab <- data.frame(
  treated = apply(pairs, 2, paste, collapse = ","),
  `x treated` = apply(pairs, 2, function(j) paste(x[j], collapse = ",")),
  `sum x` = apply(pairs, 2, function(j) sum(x[j])),
  complete = 1 / 6,
  balanced = share(Z_balanced),
  check.names = FALSE
)
knitr::kable(tab, digits = 3,
             caption = "Shares of the six assignments of two treated units. `complete_ra` is uniform; `balanced_ra(formula = ~ x)` is not.")
```

```{r}
rbind(
  complete = c(p = mean(Z_complete), treated = mean(colSums(Z_complete))),
  balanced = c(p = mean(Z_balanced), treated = mean(colSums(Z_balanced)))
)
```

The cube walk reweights the six vertices. It does not put extra mass on the pair-sums nearest 6 (rows 3 and 4). With four units and two constraints most of the work is landing, which may drop a column of $X$. Section 6 walks through those steps.

# 3. Three C++ kernels

R dispatches on the call. Clusters are collapsed first; the same three entries then run at cluster level.

```{r, echo = FALSE}
knitr::kable(
  data.frame(
    Call = c(
      "`balanced_ra(prob_unit = p)` or `balanced_ra(blocks = b)`",
      "`balanced_ra(prob_unit_each = P)`",
      "`balanced_ra(formula = ~ x)`"
    ),
    `C++` = c("`cube_two_arm_cpp`", "`cube_multi_cpp`", "`cube_on_x_cpp`"),
    Does = c(
      "Two-arm counts; leftover pairing if `blocks`",
      "Three or more arms",
      "Linear targets on a model matrix $X$"
    ),
    Paper = c(
      "Deville and Tillé (1998), pivotal method",
      "Deville and Tillé (2004), cube; Chauvet and Tillé (2006), window",
      "Deville and Tillé (2004), cube; Chauvet and Tillé (2006), window"
    ),
    check.names = FALSE
  ),
  caption = "Which C++ kernel a call uses. `formula` plus `prob_unit_each` is not implemented."
)
```

The three are not interchangeable. A general cube could mimic the pivot when $X$ is only the intercept, and the multi-arm walk on two columns is again the pivot. What the two-arm function adds is leftover pairing: two-arm blocked counts stay tight *inside each block and overall*. Generic landing drops a constraint and would lose one of those. Cube-on-X does not take `blocks`. Multi-arm plus `formula` is a further generalization and is not implemented.

```{r, echo=FALSE}
along <- function(x1, y1, x2, y2, t) {
  data.frame(x = x1 + t * (x2 - x1), y = y1 + t * (y2 - y1))
}
fmt_u <- function(u) {
  lab <- paste0("u = ", ifelse(u > 0, "+", ""), u)
  lab <- gsub("u = \\+0", "u = 0", lab)
  gsub("u = -", "u = \u2212", lab)
}
fmt_abs <- function(x) {
  # Fractions the vignette already uses; leave tenths (0.2, 0.6, 0.4) as decimals.
  pairs <- list(
    "1/20" = 1 / 20, "1/12" = 1 / 12, "1/6" = 1 / 6,
    "1/4" = 1 / 4, "1/3" = 1 / 3, "5/12" = 5 / 12,
    "1/2" = 1 / 2, "2/3" = 2 / 3
  )
  for (nm in names(pairs)) {
    if (abs(x - pairs[[nm]]) < 1e-8) return(nm)
  }
  sub("\\.?0+$", "", sprintf("%.2f", x))
}
fmt_signed <- function(x) {
  paste0(if (x > 0) "+" else "\u2212", fmt_abs(abs(x)))
}
fmt_delta_note <- function(dplus, dminus) {
  paste0("\u03B4+ = ", fmt_abs(dplus),
         "      \u03B4\u2212 = ", fmt_abs(dminus))
}
# Largest plus/minus steps with z <- z + δ u, hitting 0 or 1.
cube_step_sizes <- function(z, u) {
  dplus <- Inf
  dminus <- Inf
  for (i in seq_along(z)) {
    if (abs(u[i]) < 1e-12) next
    if (u[i] > 0) {
      dplus <- min(dplus, (1 - z[i]) / u[i])
      dminus <- min(dminus, z[i] / u[i])
    } else {
      dplus <- min(dplus, z[i] / (-u[i]))
      dminus <- min(dminus, (1 - z[i]) / (-u[i]))
    }
  }
  list(dplus = dplus, dminus = dminus)
}
walk_step_sizes <- function(P, cu, ca) {
  dplus <- Inf
  dminus <- Inf
  for (e in seq_along(cu)) {
    z <- P[cu[e], ca[e]]
    if ((e - 1) %% 2 == 0) {
      dplus <- min(dplus, 1 - z); dminus <- min(dminus, z)
    } else {
      dplus <- min(dplus, z); dminus <- min(dminus, 1 - z)
    }
  }
  list(dplus = dplus, dminus = dminus)
}
# mass[i, j] is the plus-step transfer on that cell (δ+ u). Arrows follow its sign.
cube_net <- function(P, mass = NULL, xval = NULL, zlab = NULL, ulab = NULL,
                     arm_names = NULL, dplus = NULL, dminus = NULL) {
  n <- nrow(P)
  k <- ncol(P)
  if (is.null(mass)) mass <- matrix(0, n, k)
  if (is.null(arm_names)) {
    arm_names <- if (k == 2L) c("Control", "Treat") else as.character(seq_len(k))
  }
  frac <- P > 1e-8 & P < 1 - 1e-8
  open <- rowSums(frac) > 0
  if (is.null(zlab) && k == 2L) {
    z <- P[, 2L]
    zlab <- ifelse(open, sprintf("z = %.2f", z),
                   ifelse(z >= 1 - 1e-8, "z = 1", "z = 0"))
  }
  units <- data.frame(
    i = seq_len(n), x = seq_len(n), y = 0,
    open = open, kind = ifelse(open, "open", "settled"),
    zlab = if (is.null(zlab)) "" else zlab,
    ulab = if (is.null(ulab)) "" else ulab,
    xlab = if (is.null(xval)) "" else paste0("x = ", xval),
    stringsAsFactors = FALSE
  )
  pad <- if (n <= 2L) 0 else 0.45
  arms <- data.frame(
    name = arm_names,
    x = if (k == 1L) mean(units$x) else seq(min(units$x) + pad,
                                            max(units$x) - pad,
                                            length.out = k),
    y = 2.22
  )
  grey <- data.frame(x = numeric(0), y = numeric(0),
                     xend = numeric(0), yend = numeric(0))
  arr <- data.frame(x = numeric(0), y = numeric(0),
                    xend = numeric(0), yend = numeric(0),
                    lab = character(0), lx = numeric(0), ly = numeric(0))
  for (i in seq_len(n)) {
    for (j in seq_len(k)) {
      if (!frac[i, j]) next
      ux <- units$x[i]; uy <- 0
      ax <- arms$x[j]; ay <- arms$y[1]
      mval <- mass[i, j]
      if (abs(mval) < 1e-12) {
        grey <- rbind(grey, data.frame(x = ux, y = uy, xend = ax, yend = ay))
      } else {
        if (mval > 0) {
          a <- along(ux, uy, ax, ay, 0.12); b <- along(ux, uy, ax, ay, 0.88)
        } else {
          a <- along(ax, ay, ux, uy, 0.12); b <- along(ax, ay, ux, uy, 0.88)
        }
        mid <- along(a$x, a$y, b$x, b$y, if (mval > 0) 0.36 else 0.64)
        arr <- rbind(arr, data.frame(
          x = a$x, y = a$y, xend = b$x, yend = b$y,
          lab = fmt_signed(mval),
          lx = mid$x, ly = mid$y
        ))
      }
    }
  }
  has_u <- !is.null(ulab)
  has_z <- !is.null(zlab)
  has_x <- !is.null(xval)
  has_d <- !is.null(dplus) && !is.null(dminus)
  y_z <- -0.28
  y_x <- if (has_z) -0.48 else -0.28
  y_u <- -0.28 - 0.20 * (has_z + has_x)
  y_as <- y_u - 0.22
  y_lo <- y_as - 0.28
  y_hi <- if (has_d) 3.08 else 2.72
  g <- ggplot()
  if (nrow(grey) > 0) {
    g <- g + geom_segment(
      data = grey, aes(x = x, y = y, xend = xend, yend = yend),
      colour = "grey70", linewidth = 0.45
    )
  }
  if (nrow(arr) > 0) {
    g <- g + geom_segment(
      data = arr, aes(x = x, y = y, xend = xend, yend = yend),
      colour = "grey20", linewidth = 0.75,
      arrow = arrow(length = grid::unit(0.15, "cm"), type = "closed")
    ) + geom_label(
      data = arr, aes(x = lx, y = ly, label = lab),
      size = 2.9, fontface = "bold", linewidth = 0,
      fill = "white", colour = "grey15",
      label.padding = grid::unit(0.08, "lines")
    )
  }
  g <- g +
    geom_point(data = units, aes(x, y, fill = kind),
               shape = 21, size = 7.6, colour = "grey20", stroke = 0.7) +
    geom_text(data = units, aes(x, y, label = i),
              size = 3.3, fontface = "bold", colour = "grey20") +
    geom_point(data = arms, aes(x, y),
               shape = 21, size = 8.6, fill = "grey20", colour = "grey20") +
    geom_text(data = arms, aes(x, y = y + 0.30, label = name),
              size = 3.4, fontface = "bold", colour = "grey20", vjust = 0)
  if (has_d) {
    g <- g + annotate(
      "text", x = mean(units$x), y = 2.88,
      label = fmt_delta_note(dplus, dminus),
      size = 3.3, colour = "grey20", fontface = "bold"
    )
  }
  if (has_z) {
    g <- g + geom_text(
      data = units, aes(x, y = y_z, label = zlab, colour = kind),
      size = 3.0, vjust = 1
    )
  }
  if (has_x) {
    g <- g + geom_text(
      data = units, aes(x, y = y_x, label = xlab, colour = kind),
      size = 3.0, vjust = 1
    )
  }
  if (has_u) {
    g <- g + geom_text(
      data = units, aes(x, y = y_u, label = ulab, colour = kind),
      size = 3.1, fontface = "bold", vjust = 1
    )
  }
  settled <- subset(units, !open)
  if (nrow(settled) > 0) {
    if (k == 2L) {
      settled$slab <- "assigned"
    } else {
      hit <- max.col(P[settled$i, , drop = FALSE], ties.method = "first")
      settled$slab <- paste0("arm ", arm_names[hit])
    }
    g <- g + geom_text(
      data = settled, aes(x, y = y_as, label = slab),
      size = 3.0, colour = "grey45", vjust = 1
    )
  }
  g +
    scale_fill_manual(values = c(open = "white", settled = "grey78"),
                      guide = "none") +
    scale_colour_manual(values = c(open = "grey20", settled = "grey55"),
                        guide = "none") +
    coord_cartesian(xlim = c(min(units$x) - 0.55, max(units$x) + 0.55),
                    ylim = c(y_lo, y_hi), expand = FALSE, clip = "off") +
    theme_void(base_size = 11) +
    theme(plot.margin = margin(10, 10, 8, 8),
          plot.background = element_rect(fill = "white", colour = NA),
          panel.background = element_rect(fill = "white", colour = NA))
}
# Two-arm: plus step is z <- z + δ+ u. Control changes by −δ+ u.
cube_bip <- function(z, u, xval = NULL, zlab = NULL, ulab = NULL,
                     dplus = NULL, dminus = NULL) {
  n <- length(z)
  P <- cbind(1 - z, z)
  ss <- cube_step_sizes(z, u)
  if (is.null(dplus)) dplus <- ss$dplus
  if (is.null(dminus)) dminus <- ss$dminus
  mass <- cbind(-dplus * u, dplus * u)
  if (is.null(ulab)) ulab <- fmt_u(u)
  cube_net(P, mass, xval = xval, zlab = zlab, ulab = ulab,
           dplus = dplus, dminus = dminus)
}
# Multi-arm: alternating kernel on (cu, ca); arrows are the plus transfer.
cube_walk <- function(P, cu, ca, arm_names = NULL,
                      dplus = NULL, dminus = NULL) {
  ss <- walk_step_sizes(P, cu, ca)
  if (is.null(dplus)) dplus <- ss$dplus
  if (is.null(dminus)) dminus <- ss$dminus
  mass <- matrix(0, nrow(P), ncol(P))
  for (e in seq_along(cu)) {
    s <- if ((e - 1) %% 2 == 0) 1 else -1
    mass[cu[e], ca[e]] <- s * dplus
  }
  cube_net(P, mass, arm_names = arm_names, dplus = dplus, dminus = dminus)
}
```

# 4. Two-arm pivot

`cube_two_arm_cpp` is the [pivotal method](https://doi.org/10.1093/biomet/85.1.89) of Deville and Tillé (1998). The state is a vector $z$ of treatment probabilities. The kernel comes from the count: an admissible $u$ on two open units is $(1,-1)$---intuitively whatever probability mass is added to one is subtracted from the other. Transferring mass leaves $z_i+z_j$ unchanged. The randomization is a fair bet between the two maximal steps that first hit 0 or 1. 

The figures are bipartite networks. Units sit on the bottom; arms sit on top. A grey edge is a still-fractional cell, not yet being  considered for assignment. Arrows show the plus-step transfer $\delta_+ u$ on that edge (just $\pm\delta_+$ when $|u_i|=1$). The value $z$, written under a unit, is the Treat mass. Control mass is $1-z$. 

## One draw: four units, target 2.3

We illustrate the logic of the pivot method for a case with four units and heterogeneous probabilities that do not sum to an integer.

Specifically, imagine $p = (0.2, 0.6, 0.7, 0.8)$. These probabilities sum to 2.3, and so every draw should treat 2 or 3 units. 

```{r}
p_piv <- c(0.2, 0.6, 0.7, 0.8)
sum(p_piv)
balanced_ra(prob_unit = p_piv)
```

The visit order will be $1,2,3,4$. In the illustration we assume the plus step is taken each time, though in practice this is random. At the end there is a residual (0.3) and the last assignment is via Bernoulli.

The first pair consists of units 1 and 2. These have a combined mass of $0.8$. The admissible "directions of movement" that keep the sum of probabilities constant is $u=(1,-1,0,0)$. The plus coefficient is $\delta_+=\min(1-0.2,0.6)=0.6$, which assigns unit 2 to control (since $u\delta_+ = (.6, -.6)$ which would yield new position $z' = (0.2, 0.6) + (0.6, -0.6) = (0.8, 0)$). For the negative coefficient we have $\delta_-=\min(0.2,0.4)=0.2$ which assigns unit 1 to control (since the new position would be $z' = (0.2, 0.6) - (0.2, -0.2) = (0, .8)$). We will choose between these two directions, selecting plus with probability $0.2/0.8=1/4$. Note that in this case the question is which gets assigned to control, not which gets assigned to treatment, as it is possible that neither will get assigned to treatment. 

```{r, echo=FALSE, fig.width=6.4, fig.height=4.3, fig.cap="Start. Kernel on units 1 and 2. Arrows are $\\delta_+ u$. $\\delta_+ = 0.6$ assigns unit 2 to control; $\\delta_- = 0.2$ assigns unit 1 to control."}
cube_bip(
  z = p_piv, u = c(1, -1, 0, 0),
  ulab = c("u = +1", "u = \u22121", "u = 0", "u = 0")
)
```

Assume we happened to select the plus coefficient. Then our new $z$ is $z=(0.8,0,0.7,0.8)$. Unit 1 is still open, at $0.8$, and it now pairs with unit 3. And we go again. This time  $\delta_+=0.2$ would send unit 1 to 1; $\delta_-=0.3$ would send unit 3 to 1. We choose randomly between these, selecting plus with  probability $0.3/0.5=3/5$.

```{r, echo=FALSE, fig.width=6.4, fig.height=4.3, fig.cap="After the first plus step. Unit 2 is assigned. Next pair: units 1 and 3. Arrows are $\\delta_+ u$."}
cube_bip(
  z = c(0.8, 0, 0.7, 0.8), u = c(1, 0, -1, 0),
  ulab = c("u = +1", "u = 0", "u = \u22121", "u = 0")
)
```

Lets imagine again that plus was selected. Then the new $z$ vector is  $z=(1,0,0.5,0.8)$. Units 3 and 4 now form a pair. We have possibilities $\delta_+=0.5$, which sends unit 3 to 1 and unit 4 to $0.3$; and $\delta_-=0.2$ which sends unit 4 to 1. Plus has probability $0.2/0.7=2/7$.

```{r, echo=FALSE, fig.width=6.4, fig.height=4.3, fig.cap="After the second plus step. Units 3 and 4 remain. $\\delta_+ = 0.5$ assigns unit 3 (leftover $z_4 = 0.3$); $\\delta_- = 0.2$ assigns unit 4."}
cube_bip(
  z = c(1, 0, 0.5, 0.8), u = c(0, 0, 1, -1),
  ulab = c("u = 0", "u = 0", "u = +1", "u = \u22121")
)
```

Assume again we randomly selected plus. Then $z=(1,0,1,0.3)$. One leftover remains, so there is no new pairing possible. We now randomly assign unit 4 with probability $0.3$. The result is that we treat either two or three units. 

## Leftover pairing across blocks

With `blocks`, the same pass runs inside each block and leaves at most one fractional unit per block. A second pass then pairs those leftovers as one group, so the *overall* count stays tight as well. Independent Bernoulli on each block leftover would have kept each block tight and let the overall total wander.


## C++: `cube_pivot_pass`

We see these same steps now in the `C++` function: 

```cpp
static void cube_pivot_pass(NumericVector& z, const std::vector<int>& seq,
                            const int* blk, int nb, double tol) {
  std::vector<int> open(nb + 1, -1);
  for (size_t t = 0; t < seq.size(); t++) {
    int j = seq[t];
    // Step 1: skip assigned.
    if (z[j] <= tol || z[j] >= 1.0 - tol) continue;
    int bl = blk[j];
    // Step 2: hold one open unit per block.
    if (open[bl] < 0) { open[bl] = j; continue; }

    // Step 3: kernel pair (z_i + z_j preserved).
    int i = open[bl];
    // Step 4: largest d+ and d-.
    double du = std::min(1.0 - z[i], z[j]);
    double dd = std::min(z[i], 1.0 - z[j]);
    // Step 5: fair bet, then transfer.
    if (unif_rand() < dd / (du + dd)) { z[i] += du; z[j] -= du; }
    else                              { z[i] -= dd; z[j] += dd; }

    // Step 6: housekeeping of open[].
    int keep = (z[i] > tol && z[i] < 1.0 - tol) ? i : j;
    open[bl] = (z[keep] > tol && z[keep] < 1.0 - tol) ? keep : -1;
  }
}
```

| Step | What the line does |
|:----:|--------------------|
| 1 | Units already at 0 or 1 are skipped. |
| 2 | The first still-fractional unit in a block is held. |
| 3 | A second fractional unit in that block is a pair. The kernel is $(1,-1)$, so $z_i+z_j$ is unchanged. |
| 4 | $\delta_+$ and $\delta_-$ are the largest steps that hit 0 or 1. |
| 5 | A fair bet: plus with probability $\delta_-/(\delta_++\delta_-)$, then the transfer. |
| 6 | The unit that is still fractional stays open; otherwise that block's slot is empty. |

`cube_two_arm_cpp` then pairs leftovers as if they shared a block, and Bernoulli-rounds any singleton:

```cpp
  // Step 1: collect leftovers.
  std::vector<int> left;
  for (int t = 0; t < n; t++) {
    int j = seq[t];
    if (z[j] > tol && z[j] < 1.0 - tol) left.push_back(j);
  }
  if (left.size() > 1) {
    // Step 2: fake a single block.
    std::vector<int> one(n, 1);
    // Step 3: same pivot pass (overall count stays tight).
    cube_pivot_pass(z, left, one.data(), 1, tol);
  }
  // Step 4: Bernoulli any singleton leftover.
```

# 5. Multi-arm cube

Consider now the cube method. 

`cube_multi_cpp` is the [cube method](https://doi.org/10.1093/biomet/91.4.893) of Deville and Tillé (2004) on the transportation polytope: an $N\times k$ matrix, row sums 1, column sums $t$. [Chauvet and Tillé (2006)](https://doi.org/10.1007/s00180-006-0250-2) supply the window of $k$ units so each step is linear in $N$.

A cycle in the bipartite graph of units and arms is flight: every row total and every column total is unchanged. A path is landing: only the two arms at the ends move. With two arms every maximal walk is a 4-cycle (the pivot). With three or more arms a window of $k$ still-fractional units forces a cycle until too few remain.

Flight stays inside each block. Each block is then landed on its own. Overall counts can wander when $k>2$ and there are blocks. Two-arm leftover pairing does not extend: a block can keep several leftovers, and coupling them globally can push a within-block arm count more than one away from its target.

The figures use the same network as Section 4: units on the bottom, arms on top, grey edges for still-fractional cells, arrows labeled with the plus-step mass $\delta_+ u$. Cell masses stay in the printed matrix, not on the edges. $\delta_+$ and $\delta_-$ sit on the figure.

## One draw: four units, three arms

```{r}
P3 <- rbind(
  c(0.2, 0.4, 0.4),
  c(0.4, 0.3, 0.3),
  c(0.6, 0.2, 0.2),
  c(0.8, 0.1, 0.1)
)
P3
colSums(P3)
balanced_ra(prob_unit_each = P3, conditions = 1:3)
```

Column targets are $2$, $1$, and $1$. The first window is units 1, 2, and 3. A cycle on that window preserves every total. The path below takes plus, then plus; unit 1 fills; later cycles bring in unit 4; the last move is a 4-cycle, not a path. Counts end at $(2,1,1)$.

```{r}
cube_move <- function(Z, cu, ca, u) {
  m <- length(cu)
  dplus <- Inf
  dminus <- Inf
  for (e in seq_len(m)) {
    z <- Z[cu[e], ca[e]]
    if ((e - 1) %% 2 == 0) {
      dplus <- min(dplus, 1 - z); dminus <- min(dminus, z)
    } else {
      dplus <- min(dplus, z); dminus <- min(dminus, 1 - z)
    }
  }
  up <- u < dminus / (dplus + dminus)
  for (e in seq_len(m)) {
    s <- if ((e - 1) %% 2 == 0) 1 else -1
    Z[cu[e], ca[e]] <- Z[cu[e], ca[e]] + if (up) s * dplus else -s * dminus
    if (Z[cu[e], ca[e]] < 1e-12) Z[cu[e], ca[e]] <- 0
    if (Z[cu[e], ca[e]] > 1 - 1e-12) Z[cu[e], ca[e]] <- 1
  }
  Z
}
Z3 <- P3
S3 <- list(start = Z3)
Z3 <- cube_move(Z3, c(1, 2, 2, 1), c(1, 1, 2, 2), 0.20)
S3$after_first_cycle <- Z3
Z3 <- cube_move(Z3, c(1, 3, 3, 2, 2, 1), c(1, 1, 2, 2, 3, 3), 0.20)
Z3 <- cube_move(Z3, c(3, 3, 4, 4), c(2, 1, 1, 2), 0.50)
Z3 <- cube_move(Z3, c(2, 3, 3, 4, 4, 2), c(2, 2, 1, 1, 3, 3), 0.40)
Z3 <- cube_move(Z3, c(2, 3, 3, 2), c(2, 2, 3, 3), 0.10)
S3$end <- Z3
```

```{r, echo=FALSE, fig.width=6.4, fig.height=4.1, fig.cap="Start. Grey edges are still-fractional cells. Arrows are $\\delta_+ u$ on the first cycle (units 1--2, arms 1--2)."}
cube_walk(P3, c(1, 2, 2, 1), c(1, 1, 2, 2), arm_names = c("1", "2", "3"))
```

```{r, echo=FALSE, fig.width=6.4, fig.height=4.1, fig.cap="After the first plus cycle. Row and column totals are unchanged. Arrows are $\\delta_+ u$ on the next cycle, which fills unit 1."}
cube_walk(S3$after_first_cycle, c(1, 3, 3, 2, 2, 1), c(1, 1, 2, 2, 3, 3),
          arm_names = c("1", "2", "3"))
```

```{r, echo=FALSE, fig.width=6.4, fig.height=3.8, fig.cap="End of the walk. Arm counts (2, 1, 1). Every remaining cell is 0 or 1."}
cube_net(S3$end, arm_names = c("1", "2", "3"))
```

```{r}
lapply(S3, function(M) round(M, 2))
```

## C++: `cube_move` and per-block landing

The fair bet along a walk is the same as in Section 4, with alternating sign so consecutive cells that share a node cancel.

```cpp
static void cube_move(std::vector<double>& Z, int n,
                      const std::vector<int>& cu, const std::vector<int>& ca,
                      double tol) {
  int m = cu.size();
  double dplus = R_PosInf, dminus = R_PosInf;
  // Step 1: largest d+ and d- (alternating sign).
  for (int e = 0; e < m; e++) {
    double z = Z[cu[e] + (size_t) ca[e] * n];
    if (e % 2 == 0) { dplus = std::min(dplus, 1.0 - z); dminus = std::min(dminus, z); }
    else            { dplus = std::min(dplus, z); dminus = std::min(dminus, 1.0 - z); }
  }
  if (!R_FINITE(dplus + dminus) || dplus + dminus <= 0) return;

  // Step 2: fair bet.
  bool up = unif_rand() < dminus / (dplus + dminus);
  // Step 3: apply the transfer.
  for (int e = 0; e < m; e++) {
    size_t ix = cu[e] + (size_t) ca[e] * n;
    double s = (e % 2 == 0) ? 1.0 : -1.0;
    Z[ix] += up ? s * dplus : -s * dminus;
    if (Z[ix] < tol) Z[ix] = 0.0;
    if (Z[ix] > 1.0 - tol) Z[ix] = 1.0;
  }
}
```

| Step | What the line does |
|:----:|--------------------|
| 1 | Walk the cycle with alternating sign. $\delta_+$ and $\delta_-$ are the largest steps that hit 0 or 1 on any cell of the walk. |
| 2 | A fair bet: plus with probability $\delta_-/(\delta_++\delta_-)$. |
| 3 | Add $+\delta_+$ or $-\delta_-$ times the sign of each edge. Row totals and interior column totals do not move. |

`cube_process` keeps a Chauvet--Tillé window of $k$ units with at least two fractional cells. `cube_multi_cpp` calls it once per block with paths allowed (landing):

```cpp
  // Two-arm leftover coupling does not extend. Each block is landed on
  // its own. Overall tightness may slip when several remainders land
  // the same way.
  for (int bl = 1; bl <= nb; bl++)
    cube_process(Z, n, k, bu[bl], tol, ws, true);
```

# 6. Cube-on-X

`cube_on_x_cpp` is the cube on a balancing matrix $X$, from `formula`. Two arms only. No `blocks`. The intercept in `~ x` is the count constraint; the $x$ column is the treated total of $x$. A flight step may move only in $\ker(X')$: $X'u=0$. The implementation uses a Chauvet--Tillé window of $q+1$ units, so with two columns the window has three units and a one-dimensional kernel. Landing decrements $q$ (drops the last column first) until a kernel vector exists.

The 4-cycle $u=(1,-1)$ on two units preserves the count and *not* the $x$-total, unless those units share an $x$. That is why this is not the pivot.

Units are sorted by the first non-intercept column of $X$. A coin flip then reverses that order. The walk below uses increasing $x$, no reverse.

## One draw: $x=(1,2,3,6)$, $p_i=1/2$

Start at $z=(1/2,1/2,1/2,1/2)$. The window is units 1, 2, and 3. The kernel direction (up to scale) is $u=(1,-2,1,0)$: $1-2+1=0$ and $1\cdot 1+2\cdot(-2)+3\cdot 1=0$. Both maximal steps have size $1/4$. Plus assigns unit 2 to control; minus assigns unit 2 to treatment.

```{r, echo=FALSE, fig.width=6.4, fig.height=4.7, fig.cap="Start. Window of three units. Kernel $u = (1, -2, 1, 0)$. Arrows are $\\delta_+ u$; unit 2 moves twice as far. $\\delta_\\pm = 1/4$."}
cube_bip(
  z = c(0.5, 0.5, 0.5, 0.5),
  xval = c(1, 2, 3, 6),
  u = c(1, -2, 1, 0),
  ulab = c("u = +1", "u = \u22122", "u = +1", "u = 0")
)
```

Assume plus. Then $z=(0.75,0,0.75,0.5)$. Sum $z$ is still 2; the treated $x$-total is still 6. The next window is the three open units. An admissible $u$ is $(3,0,-5,2)$: $3-5+2=0$ and $1\cdot 3+3\cdot(-5)+6\cdot 2=0$. $\delta=1/12$ assigns unit 1 to treatment; $\delta=-1/20$ assigns unit 3 to treatment. Plus has probability $3/8$. Neither of these two steps sends a unit to 0.

```{r, echo=FALSE, fig.width=6.4, fig.height=4.7, fig.cap="After the first plus step. Unit 2 is assigned. Arrows are $\\delta_+ u$ with $u = (3, 0, -5, 2)$. $\\delta_+ = 1/12$; $\\delta_- = 1/20$."}
cube_bip(
  z = c(0.75, 0, 0.75, 0.5),
  xval = c(1, 2, 3, 6),
  u = c(3, 0, -5, 2),
  ulab = c("u = +3", "u = 0", "u = \u22125", "u = +2")
)
```

Assume $\delta=1/12$. Then $z=(1,0,1/3,2/3)$. Two open units and two constraints: the kernel is $\{0\}$. A count-preserving $u_3=-u_4$ moves the $x$-total. The algorithm drops $x$.

```{r, echo=FALSE, fig.width=6.4, fig.height=4.7, fig.cap="After $\\delta = 1/12$. No kernel remains on both columns, so $x$ is dropped. Arrows are the count-only landing move $\\delta_+ u$. $\\delta_+ = 2/3$; $\\delta_- = 1/3$."}
cube_bip(
  z = c(1, 0, 1 / 3, 2 / 3),
  xval = c(1, 2, 3, 6),
  u = c(0, 0, 1, -1),
  zlab = c("z = 1", "z = 0", "z = 1/3", "z = 2/3"),
  ulab = c("u = 0", "u = 0", "u = +1", "u = \u22121")
)
```

What remains is the count. $u=(0,0,1,-1)$. $\delta=2/3$ assigns unit 3 to treatment and unit 4 to control (treated $x$-total $4$). $\delta=-1/3$ does the reverse (total $7$). Neither hits the target $6$. That is landing.

## C++: window, kernel, drop a column

```cpp
    int w = std::min(nf, q_use + 1);
    // kernel of X on that window, then z <- z ± δ u
    bool moved = try_window(z, W, Xs, n, q_use, A, u, tol);
    if (!moved) {
      q_use--;   // drop last column of X (x before the intercept)
      continue;
    }
```

`try_window` finds $u\neq 0$ in $\ker(X_{\mathrm{window}}')$ and takes the same fair bet as the pivot, now on every coordinate with $u_i\neq 0$. If the small window has empty kernel it tries all remaining units before dropping a column.

# 7. What this implementation guarantees

Every unit receives exactly one condition. Each unit's probability of each condition equals the supplied $\pi_{ij}$. Each step is a martingale, so $E[Z]=\pi$.

Condition counts are the floor or the ceiling of the target at a stated level:

- without `blocks`, overall counts are tight;
- with `blocks`, within-block counts are always tight;
- with `blocks` and two arms, overall counts are tight as well (leftover pairing);
- with `blocks` and three or more arms, overall counts can wander;
- with `clusters`, the tight counts are counts of clusters;
- with `formula`, first-order inclusion probabilities remain exact; covariate totals are as close as landing allows.

Deville and Tillé (2004) supply the martingale and the flight/landing split. They do not, by themselves, say that a blocked two-arm design is tight at both levels. That is the leftover pass.

# 8. Examples

## A race with unequal chances

Four contestants have winning probabilities 0.50, 0.30, 0.15, and 0.05. The probabilities sum to 1, so exactly one contestant should win each time. Simple random assignment gets the chances right and sometimes produces two winners or none. Complete random assignment produces one winner and equal chances.

```{r}
chances <- c(0.5, 0.3, 0.15, 0.05)
set.seed(1)
which(balanced_ra(prob_unit = chances) == 1)
winners <- replicate(2000, which(balanced_ra(prob_unit = chances) == 1))
table(winners) / 2000
```

The shares sit near the supplied chances, and every draw has a single winner.

## Random probabilities, one treated arm

Eighty units, five blocks of uneven size, probabilities drawn from the unit interval. Unit-level means should track $p$.

```{r}
set.seed(3)
n <- 80
p <- runif(n)
blocks <- sample(1:5, n, replace = TRUE, prob = 1:5)
reps <- replicate(1500, balanced_ra(prob_unit = p, blocks = blocks))
share <- rowMeans(reps)
ggplot(data.frame(p, share), aes(p, share)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40") +
  geom_point(size = 1.5) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
  labs(x = "supplied probability", y = "share treated") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

## Three arms, a tight $2 \times 3$ matrix

Arm 2 has expected count $1.13$, so a tight assignment gives that arm 1 or 2 units every time.

```{r}
P <- cbind(c(0.15, 0.47), c(0.65, 0.48), c(0.20, 0.05))
P
colSums(P)
set.seed(4)
table(replicate(2000, balanced_ra(prob_unit_each = P, conditions = 1:3))[1, ])
table(colSums(replicate(2000,
  balanced_ra(prob_unit_each = P, conditions = 1:3) == 2)))
```

## Clusters, and blocks with clusters

Whole clusters move together. Tight counts are cluster counts.

```{r}
set.seed(8)
clusters <- rep(1:6, times = c(3, 1, 4, 2, 5, 3))
p_cluster <- c(0.2, 0.4, 0.6, 0.8, 0.5, 0.5)
z <- balanced_ra(prob_unit = p_cluster[clusters], clusters = clusters)
table(clusters, z)
```

```{r}
set.seed(9)
blocks <- ifelse(clusters <= 3, "east", "west")
z <- balanced_ra(prob = 0.5, clusters = clusters, blocks = blocks)
table(blocks, z)
```

## Interface refusals

Rows of `prob_unit_each` must sum to 1. There is no implicit residual condition 0.

```{r, error=TRUE}
balanced_ra(prob_unit_each = cbind(c(0.25, 0.35, 0.40),
                                  c(0.47, 0.48, 0.05)))
```

Probabilities outside $[0,1]$ are refused.

```{r, error=TRUE}
balanced_ra(prob_unit = c(1.2, 0.8, 2.1, 0.9))
```

A single shared probability does not imply $N$: with `prob` alone, supply `N`, `blocks`, or `clusters`.

```{r, error=TRUE}
balanced_ra(prob = 0.5)
```

# 9. Validation

Replicates are 2,000. What is shown is the largest absolute gap between an empirical share and the supplied probability, and the table of realized counts.

## Unequal probabilities, no blocks

```{r}
set.seed(11)
p <- c(0.2, 0.4, 0.6, 0.8, 0.5, 0.5)
r <- replicate(2000, balanced_ra(prob_unit = p))
max(abs(rowMeans(r) - p))
table(colSums(r))
```

Every draw treats exactly 3.

## Two arms with blocks: overall and within-block

```{r}
set.seed(12)
blocks <- rep(1:10, each = 3)
r <- replicate(2000, balanced_ra(blocks = blocks))
table(colSums(r))
block_range <- sapply(1:10, function(b)
  range(colSums(r[blocks == b, , drop = FALSE])))
rownames(block_range) <- c("min", "max")
block_range
max(abs(rowMeans(r) - 0.5))
```

Every draw hits 15. Every block contributes 1 or 2.

## Three arms, no blocks: overall tight

```{r}
set.seed(13)
P <- cbind(c(0.15, 0.47), c(0.65, 0.48), c(0.20, 0.05))
Z <- replicate(2000, as.integer(as.character(
  balanced_ra(prob_unit_each = P, conditions = 1:3))))
count_tab <- sapply(1:3, function(j)
  table(factor(colSums(Z == j), levels = 0:2)))
colnames(count_tab) <- paste0("arm", 1:3)
count_tab
max(abs(rowMeans(Z == 1) - P[, 1]),
    abs(rowMeans(Z == 2) - P[, 2]),
    abs(rowMeans(Z == 3) - P[, 3]))
```

## Two-arm blocked districts, as a joint check

```{r}
set.seed(14)
districts <- rep(c("north", "south"), each = 3)
r <- replicate(2000, balanced_ra(blocks = districts))
north <- colSums(r[districts == "north", ])
south <- colSums(r[districts == "south", ])
table(north, south)
all(colSums(r) == 3)
```

The support is $\{(1,2),(2,1)\}$ only.

## Equal probabilities and `complete_ra()`

```{r}
set.seed(15)
table(colSums(replicate(1000, balanced_ra(4))))
table(colSums(replicate(1000, complete_ra(N = 4))))
```

# 10. Warnings / caveats

The function is experimental. The interface may change.

Counts are always tight within each block. With two arms they are tight overall as well. With three or more arms the blocks are landed separately, so the overall count can wander.

```{r}
set.seed(16)
P <- matrix(1/3, 6, 3)
blocks <- rep(1:3, each = 2)
Z <- replicate(2000, as.integer(as.character(
  balanced_ra(prob_unit_each = P, blocks = blocks, conditions = 1:3))))
table(colSums(Z == 1))
```

Arm 1's overall count is not confined to $\{1,2\}$. Hamilton's method would give each arm to exactly two of the three blocks; this implementation does not when $k>2$.

`formula` and `blocks` cannot be combined. `formula` and `prob_unit_each` cannot be combined: multi-arm cube-on-X is not implemented.

`check_inputs = FALSE` skips validation. It is for simulation loops whose probabilities have already been verified.

The takeaway is practical. Use `balanced_ra()` when realized assignment should stay tight against its targets. Trust overall tightness without blocks, and both overall and within-block tightness with two arms. Do not trust overall tightness with three or more arms and `blocks`. Do not pass `formula` with `prob_unit_each`.

# References

Deville, J.-C. and Tillé, Y. (1998). Unequal probability sampling without replacement through a splitting method. *Biometrika* 85(1), 89--101. <https://doi.org/10.1093/biomet/85.1.89>

Deville, J.-C. and Tillé, Y. (2004). Efficient balanced sampling: the cube method. *Biometrika* 91(4), 893--912. <https://doi.org/10.1093/biomet/91.4.893>

Deville, J.-C. and Tillé, Y. (2005). Variance approximation under balanced sampling. *Journal of Statistical Planning and Inference* 128(2), 569--591. <https://doi.org/10.1016/j.jspi.2003.11.011>

Chauvet, G. and Tillé, Y. (2006). A fast algorithm for balanced sampling. *Computational Statistics* 21(1), 53--62. <https://doi.org/10.1007/s00180-006-0250-2>
