---
title: "Assignment with heterogeneous probabilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assignment with heterogeneous probabilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
set.seed(20260822)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
options(digits = 3)
```

```{r}
library(randomizr)
library(ggplot2)
```

`balanced_ra()` assigns units to conditions when assignment probabilities vary from unit to unit, and it holds the number assigned to each condition at the floor or the ceiling of that condition's expected count. Three things hold on every draw: each unit receives exactly one condition; each unit's probability of each condition is the probability supplied; and the realized counts sit at the floor or the ceiling of their targets. With two arms and `blocks`, those tight counts are both within each block and overall. With three or more arms and `blocks`, tightness is within each block; the overall count is the sum of independent block landings and can move by more than one. Section 1 is a short demonstration. Section 2 gives the logic and states what the published papers guarantee versus what this implementation guarantees. Section 3 walks through the C++. Section 4 gives worked examples. Section 5 reports a concise validation. Section 6 lists the caveats.

The function is **experimental**. Declare it with `declare_ra(..., ra_type = "balanced")` or by supplying `prob_unit_each`.

# 1. Short demonstration

The two-district design is the case that complete random assignment inside each block cannot handle on its own. Two districts of three villages, three villages to treat, equal probabilities, blocked by district: the per-district target is 1.5. Each district should receive one or two treated villages, never zero and never three, and the total should be three on every draw.

```{r}
blocks <- rep(1:2, each = 3)
p <- (1:6)/12
balanced_ra(blocks = blocks, prob_unit = p)
```

Repeating the draw, the unit means sit at 0.5, each district contributes 1 or 2, and the total is always 3.

```{r}
reps <- replicate(5000, balanced_ra(blocks = blocks, prob_unit = p))

# individual assignment probabilities
cbind(target = p, average = rowMeans(reps))

# block totals
table(colSums(reps[blocks == 1, ]), colSums(reps[blocks == 2, ]))
```

`balanced_ra(4)` is the same idea with no blocks and the default probability 0.5: complete assignment of two treated out of four.

```{r}
table(balanced_ra(4))
```

The rest of the vignette explains why those counts stay tight, how the C++ does it, and where the guarantee stops.

# 2. Logic

`balanced_ra()` is cube/pivotal assignment with unit-varying probabilities, on the same interface as `complete_ra()` and `block_ra()`. For two-arm blocked designs leftover pairing keeps the counts tight within each block and overall.

The input is an $N \times k$ matrix $\pi$ of assignment probabilities. Row $i$ sums to 1. Column $j$ sums to a target $t_j = \sum_i \pi_{ij}$. The output is a 0-1 matrix $Z$ with the same row sums, so each unit receives one condition, and with $E[Z]=\pi$. The column sums of $Z$ are required to be as close to $t$ as integer arithmetic allows: each realized count $n_j$ satisfies $n_j \in \{\lfloor t_j \rfloor, \lceil t_j \rceil\}$.

One way to picture the algorithm is as a flow on the bipartite graph of units and conditions. An edge exists where a cell is still fractional. A cycle in that graph is a closed walk that alternates between units and conditions. Pushing a constant $\delta$ forward along the even edges and backward along the odd edges leaves every row total and every column total unchanged, and a large enough $\delta$ drives at least one cell to 0 or 1. Repeat until no cycle remains.

What remains is a forest. Every unit still has either no fractional cells or at least two of them: a unit with a single fractional cell would need that cell to be an integer, because its other cells are already 0 or 1 and the row sums to 1. So every leaf of the forest is a condition. A maximal path therefore runs from one condition to another. Pushing along that path preserves every row total and every interior column total; only the two conditions at the ends move. That is the landing step.

Each push is a fair bet. If the largest feasible up-step is $d^{+}$ and the largest feasible down-step is $d^{-}$, the algorithm takes the up-step with probability $d^{-}/(d^{+}+d^{-})$. The current matrix is then a martingale:

$$E[Z^{(t+1)} \mid Z^{(t)}] = Z^{(t)}.$$

Unrolling the martingale gives $E[Z]=\pi$. The inclusion probabilities are exactly the supplied probabilities, not an approximation to them.

The same degree bound that appears in dependent rounding explains the floor-or-ceiling counts. After the walk finishes, each vertex of the bipartite graph has integer degree equal to the floor or the ceiling of the fractional degree it started with. For a condition that degree is the number of units assigned to it. For a unit it is 1.

## What the papers establish

[Deville and Tillé (2004)](https://doi.org/10.1093/biomet/91.4.893) give the cube method for balanced sampling. The state is a vector in the unit cube. Balancing constraints are linear equalities — in survey sampling, that the Horvitz–Thompson totals of auxiliary variables match the known population totals. The *flight* phase moves in a direction that lies in the kernel of those constraints until a coordinate hits 0 or 1. The *landing* phase begins when no such direction remains, and it drops constraints until the state is a vertex. Each flight step is constructed to be a martingale, so first-order inclusion probabilities equal the supplied $\pi$. Landing may violate a dropped constraint. The paper does not claim that every balancing constraint survives all the way to the assignment.

This function applies that method to a particular constraint system: the transportation polytope of nonnegative $N \times k$ matrices with row sums 1 and column sums $t$. Cycle moves are the kernel directions that preserve every row and every column. Path moves are landing. Deville and Tillé (2004) supply the martingale and the flight/landing split. They do not, by themselves, say that a blocked assignment problem will be tight both within each block and overall.

[Deville and Tillé (1998)](https://doi.org/10.1093/biomet/85.1.89) give the splitting method for unequal-probability sampling without replacement. The *pivotal* method is the two-unit case: transfer mass between two units until one hits 0 or 1, with the direction chosen so that the step is a martingale and the total remains exact. With two arms the transportation graph is a forest of stars — each unit has one edge, to its own block — so every maximal path is unit–block–unit. The general cube walk collapses to the pivotal method. That is the two-arm path in the C++.

[Chauvet and Tillé (2006)](https://doi.org/10.1007/s00180-006-0250-2) give a fast flight phase whose cost is linear in the population size. The device is a working window: only as many units as there are balancing constraints need to be examined at once. In this problem the window is $k$ units. A unit that is still fractional has at least two fractional cells, so $k$ such units induce at least $2k$ edges on at most $2k$ nodes, and a graph with at least as many edges as nodes is not a forest. The window therefore always contains a cycle until too few fractional units remain. That is the three-or-more-arm path, in the form this problem takes it.

[Deville and Tillé (2005)](https://doi.org/10.1016/j.jspi.2003.11.011) approximate the variance of Horvitz–Thompson totals under balanced sampling, when joint inclusion probabilities have no closed form. This function does not implement those estimators.

## What this implementation guarantees

Every unit receives exactly one condition. Each unit's probability of each condition equals the supplied $\pi_{ij}$. Condition counts are the floor or the ceiling of the target at a stated level:

- without `blocks`, the overall counts are tight;
- with `blocks`, the within-block counts are always tight;
- with `blocks` and two arms, the overall counts are tight as well;
- with `clusters`, the tight counts are counts of clusters, not of units.

The two-arm overall-and-block result is an implementation choice, not a named theorem that "the cube always does both." After a first pivotal pass inside each block, at most one unit per block is still fractional. A second pass pairs those leftovers as a single group. Pairing preserves their total, so the overall count stays tight, and each leftover is still sent to 0 or 1, so no block moves by more than one from its target. Independent Bernoulli rounding of each leftover would have kept each block tight and let the overall count drift by up to the number of blocks.

That leftover pairing does not extend to three or more arms. A block can keep several leftover units, and landing those leftovers as one global group can push a within-block arm count more than one away from its target. The implementation therefore lands each block on its own. Within-block counts stay at the floor or the ceiling. Overall counts are the sum of those independent landings. When several block remainders fall the same way, the overall count can move by more than one. Joint floor-or-ceiling at both levels is feasible — that is Hamilton's method of apportionment, applied to the block remainders — and this implementation delivers it for two arms, not for $k>2$.

# 3. Code walk-through

The R wrapper builds the probability matrix, then hands a shuffled unit order to C++. The shuffle is there so that the input order does not matter. Two arms and three or more arms take different functions in `src/cube.cpp`. Clusters never reach those functions as unit-level rows: they are collapsed first.

```r
cube_assign <- function(P, blocks = NULL, tol = 1e-12) {
  n <- nrow(P)
  b <- if (is.null(blocks)) rep(1L, n) else as.integer(factor(blocks))
  ord <- sample.int(n)
  if (ncol(P) == 2L) {
    z <- cube_two_arm_cpp(P[, 2L], b, ord, tol)
    return(cbind(1 - z, z))
  }
  cube_multi_cpp(P, b, ord, tol)
}
```

With `clusters`, `cube_assign_clusters()` keeps one row per cluster (the first unit; probabilities are constant within a cluster by the time this is reached), assigns at that level, and expands the 0-1 rows back.

```r
cube_assign_clusters <- function(P, clusters, blocks = NULL, tol = 1e-12) {
  cl <- factor(clusters)
  first <- match(levels(cl), as.character(cl))
  Pc <- P[first, , drop = FALSE]
  bc <- if (is.null(blocks)) NULL else blocks[first]
  Zc <- cube_assign(Pc, bc, tol)
  Zc[as.integer(cl), , drop = FALSE]
}
```

The tight counts are then counts of clusters. Clusters of unequal size give a fixed number of treated clusters and a varying number of treated units, which is how `cluster_ra()` behaves.

## Two-arm pivotal pass

`cube_pivot_pass` holds one open fractional unit per block. When a second fractional unit in that block appears, the two are a pivotal pair.

```cpp
static void cube_pivot_pass(NumericVector& z, const std::vector<int>& seq,
                            const int* blk, int nb, double tol) {
  std::vector<int> open(nb + 1, -1);
  for (size_t t = 0; t < seq.size(); t++) {
    int j = seq[t];
    if (z[j] <= tol || z[j] >= 1.0 - tol) continue;
    int bl = blk[j];
    if (open[bl] < 0) { open[bl] = j; continue; }

    int i = open[bl];
    double du = std::min(1.0 - z[i], z[j]);
    double dd = std::min(z[i], 1.0 - z[j]);
    if (unif_rand() < dd / (du + dd)) { z[i] += du; z[j] -= du; }
    else                              { z[i] -= dd; z[j] += dd; }

    int keep = (z[i] > tol && z[i] < 1.0 - tol) ? i : j;
    open[bl] = (z[keep] > tol && z[keep] < 1.0 - tol) ? keep : -1;
  }
}
```

The quantities `du` and `dd` are $d^{+}$ and $d^{-}$ above. One of the two units hits 0 or 1. The survivor, if still fractional, becomes the new open unit for that block. One pass leaves at most one leftover per block.

## Two-arm leftover coupling

`cube_two_arm_cpp` then collects those leftovers and, if there are at least two, runs the same pass as if they all shared a single block.

```cpp
  if (left.size() > 1) {
    std::vector<int> one(n, 1);
    cube_pivot_pass(z, left, one.data(), 1, tol);
  }
```

The second pass preserves the sum of the leftovers, so the overall treated count stays at the floor or the ceiling of $\sum_i \pi_i$. Each leftover is still sent to 0 or 1, so each block stays within one of its target. Any single unit that remains after pairing is rounded with a Bernoulli draw at its current value. That last draw is still a martingale, and a single leftover cannot move the overall count by more than one.

### One four-unit draw

Two blocks of two. The probabilities sum to 2, so exactly two units are treated. The walk is on a four-by-two matrix: rows are units, columns are control $1-p$ and treatment $p$. The path below ends at $(1,0,0,1)$.

```{r}
p4 <- c(0.2, 0.4, 0.6, 0.8)
blocks4 <- c(1, 1, 2, 2)
c(overall = sum(p4), setNames(tapply(p4, blocks4, sum), paste0("block", 1:2)))
```

Block 1 is the two lighter units. Its target is $0.6$, so that block receives 0 or 1, never 2. Block 2 has target $1.4$, so that block receives 1 or 2. The overall total is 2. The joint distribution of the block counts can put mass only on $(0,2)$ and $(1,1)$.

Among the six assignments that treat exactly two units, five of those pairs are allowed. The sixth is $(1,1,0,0)$: both of the first two, none of the last two. A design that only tightens the overall total of 2 can draw that pattern. With these blocks it is excluded.

```{r}
pat <- function(z) paste(z, collapse = "")
allowed <- c("1010", "1001", "0110", "0101", "0011")
data.frame(
  assignment = c("1100", allowed),
  block1 = c(2, 1, 1, 1, 1, 0),
  block2 = c(0, 1, 1, 1, 1, 2),
  overall = 2,
  with_blocks = c("no", rep("yes", 5))
)
```

```{r}
zb <- replicate(2000, balanced_ra(blocks = blocks4, prob_unit = p4,
                                 check_inputs = FALSE))
zn <- replicate(2000, balanced_ra(prob_unit = p4, check_inputs = FALSE))
c(both_first_with_blocks = mean(colSums(zb[1:2, ]) == 2),
  both_first_no_blocks = mean(colSums(zn[1:2, ]) == 2))
sort(table(apply(zb, 2, pat)))
sort(table(apply(zn, 2, pat)))
```

With blocks, $(1,1,0,0)$ does not appear. Without blocks the overall total is still 2, and $(1,1,0,0)$ does appear. The first two units are never both treated once the block constraint is on.

One shuffle and one sequence of uniforms produce a 0-1 assignment. The wrapper shuffles the units. This path uses the order $1,2,3,4$. The uniforms are $0.20$, $0.50$, $0.40$. The four panels are that path. A filled cell is the chosen assignment. A grey cell with an X is settled at 0. A number is still fractional. A heavier border marks the cells that moved in that step. Panel titles name the procedure. The numbered paragraphs immediately below the figure walk through each step: which units are still open, which sampling rule is used, what the uniform selected, and which cell was filled or deleted and why.

```{r}
pivot <- function(z, i, j, u) {
  du <- min(1 - z[i], z[j])
  dd <- min(z[i], 1 - z[j])
  if (u < dd / (du + dd)) {
    z[i] <- z[i] + du
    z[j] <- z[j] - du
  } else {
    z[i] <- z[i] - dd
    z[j] <- z[j] + dd
  }
  z
}
z <- p4
S <- rbind(P = z)
z <- pivot(z, 1, 2, 0.20); S <- rbind(S, "block 1: 1 with 2" = z)
z <- pivot(z, 3, 4, 0.50); S <- rbind(S, "block 2: 3 with 4" = z)
z <- pivot(z, 1, 3, 0.40); S <- rbind(S, "leftovers: 1 with 3" = z)
colnames(S) <- paste0("z", 1:4)
```

```{r, echo = FALSE}
cube_tile_plot <- function(states, panel_labs, arm_labels, unit_labels,
                           hline = NULL, ncol = 2, text_size = 4.2) {
  long <- do.call(rbind, lapply(seq_along(states), function(s) {
    M <- states[[s]]
    prev <- if (s == 1L) NULL else states[[s - 1L]]
    d <- expand.grid(unit = seq_len(nrow(M)), arm = seq_len(ncol(M)),
                     KEEP.OUT.ATTRS = FALSE)
    d$panel <- panel_labs[[s]]
    d$value <- M[cbind(d$unit, d$arm)]
    d$changed <- if (is.null(prev)) FALSE else
      abs(d$value - prev[cbind(d$unit, d$arm)]) > 1e-8
    d
  }))
  long$panel <- factor(long$panel, levels = panel_labs)
  long$arm_lab <- factor(long$arm, levels = seq_along(arm_labels),
                         labels = arm_labels)
  long$unit_lab <- factor(unit_labels[long$unit], levels = unit_labels)
  long$kind <- ifelse(long$value >= 1 - 1e-8, "one",
               ifelse(long$value <= 1e-8, "zero", "frac"))
  long$label <- ifelse(long$kind == "one", "1",
                ifelse(long$kind == "zero", "X",
                       sprintf("%.1f", long$value)))
  long$lab_col <- ifelse(long$kind == "one", "white", "grey15")
  long$edge <- ifelse(long$changed, "moved", "same")
  g <- ggplot(long, aes(arm_lab, unit_lab)) +
    geom_tile(aes(fill = kind, linewidth = edge),
              colour = "grey25", width = 0.92, height = 0.86) +
    geom_text(aes(label = label, colour = lab_col),
              size = text_size, fontface = "bold") +
    scale_fill_manual(values = c(one = "grey20", zero = "grey80", frac = "white"),
                      guide = "none") +
    scale_colour_identity() +
    scale_linewidth_manual(values = c(same = 0.35, moved = 1.2),
                           guide = "none") +
    scale_y_discrete(limits = rev) +
    facet_wrap(~ panel, ncol = ncol) +
    labs(x = NULL, y = NULL,
         caption = "Filled = assigned (1).   Grey + X = out (0).   Number = still fractional.   Heavy border = cells that moved.") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      strip.text = element_text(face = "bold", hjust = 0, size = 9.5,
                                lineheight = 1.1),
      axis.text = element_text(colour = "grey20"),
      plot.caption = element_text(hjust = 0, size = 8, colour = "grey30"),
      panel.spacing = unit(0.8, "lines")
    )
  if (!is.null(hline)) {
    g <- g + geom_hline(yintercept = hline, linetype = 2, colour = "grey45",
                        linewidth = 0.4)
  }
  g
}
```

```{r, echo = FALSE, fig.width = 6.8, fig.height = 7.6, fig.cap = "Four states of the probability matrix on the path to (1, 0, 0, 1). Panel titles name the procedure. Numbered paragraphs below walk through each step. Filled = assigned (1). Grey cell with an X = settled at 0. Number = still fractional. Heavier border = cells that moved in that step."}
as_P <- function(z) cbind(Control = 1 - as.numeric(z), Treat = as.numeric(z))
panel_labs <- c(
  "Step 1 — start",
  "Step 2 — pivotal pair, block 1",
  "Step 3 — pivotal pair, block 2",
  "Step 4 — leftover pair\nacross blocks"
)
cube_tile_plot(
  states = lapply(seq_len(nrow(S)), function(i) as_P(S[i, ])),
  panel_labs = panel_labs,
  arm_labels = c("Control", "Treat"),
  unit_labels = c("B1 · 1", "B1 · 2", "B2 · 3", "B2 · 4"),
  hline = 2.5
)
```

1. **Step 1 — start.** All four units are still open (fractional). Treatment probabilities are $0.2$, $0.4$, $0.6$, and $0.8$. No pairing has been drawn yet. The pass stores unit 1 as block 1's open unit and waits for a second fractional unit in that block.

2. **Step 2 — pivotal pair, block 1.** Units 1 and 2 are both still open and sit in the same block, so we take a *pivotal pair*: one transfer of treated mass between them that leaves the block total at $0.6$. There are two directions. Plus moves as much as possible from unit 2 onto unit 1 (landing at $0.6$ and $0$). Minus moves as much as possible the other way (landing at $0$ and $0.6$). We take plus with probability $1/3$. Imagine we happened to choose plus. Then $0.4$ moves from unit 2 to unit 1: unit 2's treatment cell is crossed out (that was all of its mass) and unit 1 is raised to $0.6$. Unit 2 is settled as control. The first two units cannot both still be treated: their combined mass is only $0.6$, so this pair always drives one of them to $0$ before either can hit $1$. Minus would have settled unit 1 instead. Either way, one of the two is already out.

3. **Step 3 — pivotal pair, block 2.** Still open: unit 1 at $0.6$ (the leftover from block 1) and both units in block 2. Unit 3 opened the block; unit 4 pairs with it. Same kind of pair: plus lands at (unit 3 to $1$, unit 4 to $0.4$); minus lands at (unit 3 to $0.4$, unit 4 to $1$). We take plus with probability $1/3$. Imagine we happened to choose minus. Then $0.2$ moves from unit 3 to unit 4: unit 4's treatment cell fills (the gap from $0.8$ to $1$) and unit 3 is lowered to $0.4$ so the block total stays $1.4$. Unit 4 is settled as treated; unit 3 remains open.

4. **Step 4 — leftover pair across blocks.** Still open: unit 1 at $0.6$ and unit 3 at $0.4$. Units 2 and 4 are already settled. We pair the leftovers as if they shared a block, so the same step preserves their combined total of $1$ and the overall treated count stays at $2$. Plus lands at (unit 1 to $1$, unit 3 to $0$); minus lands at (unit 1 to $0$, unit 3 to $1$). We take plus with probability $0.6$. Imagine we happened to choose plus. Then $0.4$ moves from unit 3 to unit 1: unit 1 fills and unit 3 is crossed out. Nothing remains for a Bernoulli draw. If one leftover had remained, it would be rounded by a fair coin at its current $z$.

The treatment column of those four matrices is the state table.

```{r}
round(S, 2)
```

Two treated. The block counts are $(1,1)$. After the within-block pass the already-settled units contribute 1 and the leftovers sum to 1, so the second pass keeps the total at 2. Independent Bernoulli on those leftovers would have allowed totals of 1, 2, or 3. The first two units cannot both be 1: block 1's pair always zeros one of them, because their combined mass $0.6$ is less than 1. This is one path. Averaging over shuffles and the plus/minus draws recovers $p$.

## Multi-arm move

With three or more arms the state is the full matrix. `cube_move` takes a walk of cells — a cycle or a path — and applies the same fair bet.

```cpp
static void cube_move(std::vector<double>& Z, int n,
                      const std::vector<int>& cu, const std::vector<int>& ca,
                      double tol) {
  int m = cu.size();
  double dplus = R_PosInf, dminus = R_PosInf;
  for (int e = 0; e < m; e++) {
    double z = Z[cu[e] + (size_t) ca[e] * n];
    if (e % 2 == 0) { dplus = std::min(dplus, 1.0 - z); dminus = std::min(dminus, z); }
    else            { dplus = std::min(dplus, z); dminus = std::min(dminus, 1.0 - z); }
  }
  if (!R_FINITE(dplus + dminus) || dplus + dminus <= 0) return;

  bool up = unif_rand() < dminus / (dplus + dminus);
  for (int e = 0; e < m; e++) {
    size_t ix = cu[e] + (size_t) ca[e] * n;
    double s = (e % 2 == 0) ? 1.0 : -1.0;
    Z[ix] += up ? s * dplus : -s * dminus;
    if (Z[ix] < tol) Z[ix] = 0.0;
    if (Z[ix] > 1.0 - tol) Z[ix] = 1.0;
  }
}
```

Consecutive cells share a node, so alternating the sign leaves every unit's row total and every interior arm's column total untouched. A cycle has no ends, so every column total is untouched as well. A path disturbs only the two arms at the ends.

## Two-core versus path

`cube_step` builds the bipartite graph on a working set of at most $k$ units, then strips leaves. What survives is the 2-core: every remaining vertex has degree at least 2, so a walk inside it cannot dead-end and must close. Finding a cycle whenever one exists is what keeps the arm totals exact during flight. Path moves are used only when the core is empty.

```cpp
  ws.q.clear();
  for (int v = 0; v < nv; v++) if (ws.deg[v] == 1) ws.q.push_back(v);
  while (!ws.q.empty()) {
    int v = ws.q.back(); ws.q.pop_back();
    if (ws.vdead[v] || ws.deg[v] != 1) continue;
    ws.vdead[v] = 1;
    for (int t = 0; t < ws.alen[v]; t++) {
      int e = ws.adj[(size_t) v * ws.stride + t];
      if (ws.edead[e]) continue;
      ws.edead[e] = 1;
      int o = (v == ws.eu[e]) ? (w + ws.ea[e]) : ws.eu[e];
      ws.deg[v]--; ws.deg[o]--;
      if (ws.deg[o] == 1) ws.q.push_back(o);
    }
  }
```

If no core remains and paths are allowed, the walk starts at an arm of original degree 1. Every unit has degree at least 2, so every leaf is an arm, and that start gives a maximal arm-to-arm path.

## Windowed flight and per-block landing

`cube_process` is the Chauvet–Tillé window. It keeps a working set $W$ of up to $k$ units that still have at least two fractional cells, takes a step, and drops units that have settled.

```cpp
static void cube_process(std::vector<double>& Z, int n, int k,
                         const std::vector<int>& units, double tol,
                         CubeWork& ws, bool allow_path) {
  size_t ptr = 0;
  std::vector<int> W;
  W.reserve(k);
  long long guard = (long long) units.size() * k + 10;
  while (guard-- > 0) {
    while ((int) W.size() < k && ptr < units.size()) {
      int u = units[ptr++];
      if (cube_nfrac(Z, n, k, u, tol) >= 2) W.push_back(u);
    }
    if (W.empty()) break;
    if (!cube_step(Z, n, k, W, tol, ws, allow_path)) break;
    std::vector<int> keep;
    for (size_t t = 0; t < W.size(); t++) {
      if (cube_nfrac(Z, n, k, W[t], tol) >= 2) keep.push_back(W[t]);
    }
    W.swap(keep);
  }
}
```

`cube_multi_cpp` calls that process once per block with `allow_path = true`. Flight and landing both stay inside the block. There is no second pass across leftovers.

```cpp
  // Two-arm leftover coupling does not extend: a block can keep several
  // leftover units, and landing them as one global group can push a
  // block-arm count more than one away from its target. Each block is
  // therefore landed on its own. Overall tightness then follows when every
  // block target is an integer, and may slip when several remainders land
  // the same way.
  for (int bl = 1; bl <= nb; bl++)
    cube_process(Z, n, k, bu[bl], tol, ws, true);
```

That last comment is the literature-versus-code distinction in the source itself. The cube method, applied independently inside each block, does not couple overall remainders.

### One four-by-three draw

Four units, three arms. Column targets are $2$, $1$, and $1$, so a tight assignment is exactly those counts.

```{r}
P3 <- rbind(
  c(0.2, 0.4, 0.4),
  c(0.4, 0.3, 0.3),
  c(0.6, 0.2, 0.2),
  c(0.8, 0.1, 0.1)
)
P3
colSums(P3)
```

The window holds three units, so the first move is a cycle. The walk below is the unused-edge walk from `cube_step`, then the fair bet from `cube_move`. An unused-edge cycle is a closed walk on the 2-core of the bipartite graph of still-open units and arms: from a starting vertex, take the first unused allowed edge, and continue until a vertex is revisited. Because every remaining vertex has degree at least 2, that walk cannot dead-end and must close. A cycle has no ends, so every row total and every column total is unchanged. A *path* would appear only if the 2-core were empty; then the walk would run from one leftover arm to another and only those two column totals would move. Order is $1,2,3,4$. This path chooses plus, plus, plus, minus, plus. The walk ends at arms $(1,2,3,1)$. The counts are $(2,1,1)$. Panel titles name the procedure. The numbered paragraphs immediately below the figure walk through each step.

```{r}
cube_move <- function(Z, cu, ca, u) {
  m <- length(cu)
  dplus <- Inf
  dminus <- Inf
  for (e in seq_len(m)) {
    z <- Z[cu[e], ca[e]]
    if ((e - 1) %% 2 == 0) {
      dplus <- min(dplus, 1 - z)
      dminus <- min(dminus, z)
    } else {
      dplus <- min(dplus, z)
      dminus <- min(dminus, 1 - z)
    }
  }
  up <- u < dminus / (dplus + dminus)
  for (e in seq_len(m)) {
    s <- if ((e - 1) %% 2 == 0) 1 else -1
    Z[cu[e], ca[e]] <- Z[cu[e], ca[e]] + if (up) s * dplus else -s * dminus
    if (Z[cu[e], ca[e]] < 1e-12) Z[cu[e], ca[e]] <- 0
    if (Z[cu[e], ca[e]] > 1 - 1e-12) Z[cu[e], ca[e]] <- 1
  }
  Z
}
Z3 <- P3
S3 <- list(start = Z3)
Z3 <- cube_move(Z3, c(1, 2, 2, 1), c(1, 1, 2, 2), 0.20)
S3$cycle1 <- Z3
Z3 <- cube_move(Z3, c(1, 3, 3, 2, 2, 1), c(1, 1, 2, 2, 3, 3), 0.20)
S3$cycle2 <- Z3
Z3 <- cube_move(Z3, c(3, 3, 4, 4), c(2, 1, 1, 2), 0.50)
S3$cycle3 <- Z3
Z3 <- cube_move(Z3, c(2, 3, 3, 4, 4, 2), c(2, 2, 1, 1, 3, 3), 0.40)
S3$cycle4 <- Z3
Z3 <- cube_move(Z3, c(2, 3, 3, 2), c(2, 2, 3, 3), 0.10)
S3$cycle5 <- Z3
```

```{r, echo = FALSE, fig.width = 7.0, fig.height = 10.2, fig.cap = "Six states of a four-by-three cube walk to arms (1, 2, 3, 1). Panel titles name the procedure. Numbered paragraphs below walk through each step. Filled = assigned (1). Grey cell with an X = settled at 0. Number = still fractional. Heavier border = cells that moved in that step."}
panel_labs3 <- c(
  "Step 1 — start",
  "Step 2 — unused-edge cycle\nunits 1–2",
  "Step 3 — unused-edge cycle\nunit 1 fills",
  "Step 4 — unused-edge cycle\nunit 4 enters",
  "Step 5 — unused-edge cycle\nunit 4 fills",
  "Step 6 — leftover 4-cycle\nnot a path"
)
cube_tile_plot(
  states = S3,
  panel_labs = panel_labs3,
  arm_labels = c("1", "2", "3"),
  unit_labels = as.character(1:4),
  ncol = 2,
  text_size = 3.8
)
```

1. **Step 1 — start.** All twelve cells are still open. Column targets are $2$, $1$, and $1$. The working window fills with units 1, 2, and 3 — the first three units that still have at least two fractional cells. Three such units on three arms force a cycle: at least six edges on at most six nodes is not a forest. No move has been drawn yet.

2. **Step 2 — unused-edge cycle on units 1–2, arms 1–2.** Still open: all four units. The first unused edges on the 2-core close on units 1 and 2 and arms 1 and 2. That is the same fair bet as the pivotal pair, now along a cycle: even cells $(1,1)$ and $(2,2)$ go up and odd cells $(2,1)$ and $(1,2)$ go down (or the reverse). Plus lands with the odd cells at $0$; minus goes the other way. We take plus with probability $1/3$. Imagine we happened to choose plus. Then $0.4$ leaves each odd cell — that was all of their mass — and the even cells rise by $0.4$. Every row total and every column total is unchanged.

3. **Step 3 — unused-edge cycle; unit 1 fills.** Still open: unit 1 (arm 1 at $0.6$ and arm 3 at $0.4$) and units 2–4. The walk now closes on units 1, 3, and 2 and all three arms. We take plus with probability $1/3$. Imagine we happened to choose plus. Then $0.4$ fills unit 1's arm-1 cell (the gap from $0.6$ to $1$) and crosses out its arm-3 cell. Unit 1 drops from the window. Unit 4 enters.

4. **Step 4 — unused-edge cycle; unit 4's arm-2 cell hits 0.** Still open: units 2, 3, and 4. The cycle is on units 3 and 4 and arms 1 and 2. Plus takes a step of $0.1$; minus a step of $0.6$. We take plus with probability about $0.86$. Imagine we happened to choose plus. Then unit 4's arm-2 cell is crossed out (that was all of its mass). Column totals are still untouched.

5. **Step 5 — unused-edge cycle; unit 4 fills.** Still open: units 2, 3, and 4 (unit 4 still has arm 1 at $0.9$ and arm 3 at $0.1$). Plus would step $0.7$; minus $0.1$. We take plus with probability $0.125$. Imagine we happened to choose minus. Then $0.1$ fills unit 4's arm-1 cell and crosses out its arm-3 cell. Unit 4 is assigned to arm 1. Arm 1's column is now the integer $2$.

6. **Step 6 — leftover 4-cycle, not a path.** Still open: units 2 and 3 on arms 2 and 3. That graph is a 4-cycle — every vertex has degree 2 — so the 2-core is nonempty and we still take a cycle, not a landing path. Plus fills the even cells $(2,2)$ and $(3,3)$; minus does the reverse. We take plus with probability $0.2$. Imagine we happened to choose plus. Those even cells fill and the odd cells are crossed out. Arms are $(1,2,3,1)$. Counts are $(2,1,1)$. A path would appear only if the 2-core were empty and a column were still fractional. That is the two-unit design with rows $(0.5,0.3,0.2)$ and targets $1$, $0.6$, and $0.4$: two cycles, then a path on the leftover unit between the two arms that are still fractional. Here every target is an integer, so flight finishes.

The six matrices are the state list.

```{r}
lapply(S3, function(M) round(M, 2))
```

Arm counts are $(2,1,1)$. The five cycle moves preserved every total. That is flight without a landing path.

# 4. Examples

Each example below is a `balanced_ra()` design. The point of each is in the paragraph that precedes the code.

## A race with unequal chances

Four contestants have winning probabilities 0.50, 0.30, 0.15, and 0.05. The probabilities sum to 1, so exactly one contestant should win each time, and the winner's identity should track those chances. Simple random assignment gets the chances right and sometimes produces two winners or none. Complete random assignment produces one winner and equal chances.

```{r}
chances <- c(0.5, 0.3, 0.15, 0.05)
set.seed(1)
which(balanced_ra(prob_unit = chances) == 1)
winners <- replicate(2000, which(balanced_ra(prob_unit = chances) == 1))
table(winners) / 2000
```

The shares sit near the supplied chances, and every draw has a single winner.

## Two districts of three

This is the motivating blocked design from Section 1. The cross-tabulation of north and south treated counts has mass only on $(1,2)$ and $(2,1)$.

```{r}
districts <- rep(c("north", "south"), each = 3)
set.seed(2)
ns <- replicate(2000, {
  z <- balanced_ra(blocks = districts)
  c(north = sum(z[districts == "north"]),
    south = sum(z[districts == "south"]))
})
table(north = ns["north", ], south = ns["south", ])
```

No draw treats zero or three in a district. No draw treats two or four overall.

## Random probabilities, one treated arm

One hundred units, five blocks of uneven size, probabilities drawn from the unit interval. Unit-level means should track $p$; the scatter is the check that they do. Tightness of the totals is left to Section 5.

```{r}
set.seed(3)
n <- 80
p <- runif(n)
blocks <- sample(1:5, n, replace = TRUE, prob = 1:5)
reps <- replicate(1500, balanced_ra(prob_unit = p, blocks = blocks))
share <- rowMeans(reps)
ggplot(data.frame(p, share), aes(p, share)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40") +
  geom_point(size = 1.5) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
  labs(x = "supplied probability", y = "share treated") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

## Three arms, a tight $2 \times 3$ matrix

The $2 \times 3$ matrix below has arm 2 with expected count $1.13$, so a tight assignment gives that arm 1 or 2 units every time.

```{r}
P <- cbind(c(0.15, 0.47), c(0.65, 0.48), c(0.20, 0.05))
P
colSums(P)
set.seed(4)
table(replicate(2000, balanced_ra(prob_unit_each = P, conditions = 1:3))[1, ])
table(colSums(replicate(2000,
  balanced_ra(prob_unit_each = P, conditions = 1:3) == 2)))
```

Arm 2 receives 1 or 2 units on every draw. The first unit's condition shares sit near its row of $P$.

## Four blocks, four units, equal shares

A neat integer design: sixteen units, four arms, probability $1/4$, four blocks of four. Each block can receive one unit per arm, and the one-draw table does.

```{r}
set.seed(5)
blocks <- rep(1:4, each = 4)
P <- matrix(0.25, 16, 4)
z <- balanced_ra(prob_unit_each = P, blocks = blocks)
table(blocks, z)
```

## Cyclic three-by-three, tiled across two blocks

Each $3 \times 3$ tile has a cyclic support: unit 1 cannot receive arm 3, unit 2 cannot receive arm 1, unit 3 cannot receive arm 2. Two blocks of six stack four such tiles. The assignment has to respect the zeros and still land each block tightly.

```{r}
set.seed(6)
tile <- rbind(c(2/3, 1/3, 0),
              c(0,   2/3, 1/3),
              c(1/3, 0,   2/3))
P <- rbind(tile, tile, tile, tile)
blocks <- rep(1:2, each = 6)
z <- balanced_ra(prob_unit_each = P, blocks = blocks)
table(z, blocks)
```

## Twelve units, four blocks of three, shares 1/2, 1/4, 1/4

Each block of three has targets $1.5$, $0.75$, $0.75$. Tightness inside the block means arm 1 receives 1 or 2 units and arms 2 and 3 receive 0 or 1.

```{r}
set.seed(7)
P <- t(replicate(12, c(0.5, 0.25, 0.25)))
blocks <- rep(1:4, each = 3)
z <- balanced_ra(prob_unit_each = P, blocks = blocks)
table(z, blocks)
```

## Clusters, and blocks with clusters

Whole clusters move together. The probabilities must be constant inside a cluster. The counts that are held tight are cluster counts. Unequal cluster sizes therefore fix the number of treated clusters and leave the number of treated units free to move.

```{r}
set.seed(8)
clusters <- rep(1:6, times = c(3, 1, 4, 2, 5, 3))
p_cluster <- c(0.2, 0.4, 0.6, 0.8, 0.5, 0.5)
z <- balanced_ra(prob_unit = p_cluster[clusters], clusters = clusters)
table(clusters, z)
```

Six clusters, probabilities summing to 3: three clusters treated. Fifteen units, of which the treated number depends on which clusters were drawn.

Blocks and clusters combine. Every cluster must sit inside one block. The tight counts are then treated clusters within each block, and — with two arms — overall as well.

```{r}
set.seed(9)
blocks <- ifelse(clusters <= 3, "east", "west")
z <- balanced_ra(prob_unit = 0.5, clusters = clusters, blocks = blocks)
table(blocks, z)
```

## Default probability, $N$ or blocks only

`balanced_ra(4)` was in Section 1. The same default $p=0.5$ with blocks only is complete assignment inside the implied design: five blocks of three, target $1.5$ per block, overall target $7.5$.

```{r}
set.seed(10)
table(balanced_ra(blocks = rep(1:5, each = 3)))
```

## Interface refusals

Three refusals are deliberate. They are not papered over.

Rows of `prob_unit_each` must sum to 1. There is no implicit residual condition 0.

```{r, error=TRUE}
balanced_ra(prob_unit_each = cbind(c(0.25, 0.35, 0.40),
                                  c(0.47, 0.48, 0.05)))
```

Add the residual column if that is the design: `cbind(P, 1 - rowSums(P))`.

Probabilities outside $[0,1]$ are refused. Values such as $1.2$ are not treated as expected counts.

```{r, error=TRUE}
balanced_ra(prob_unit = c(1.2, 0.8, 2.1, 0.9))
```

A scalar probability does not imply $N$. Supply `N`, `blocks`, or `clusters`.

```{r, error=TRUE}
balanced_ra(prob_unit = 0.5)
```

# 5. Validation

The claims in Section 2 are per-draw properties plus a marginal. This section reports them on a short set of designs. Replicates are 2,000 throughout. What is shown is the largest absolute gap between an empirical share and the supplied probability, and the table of realized counts.

## Unequal probabilities, no blocks

Six units, $p=(0.2,0.4,0.6,0.8,0.5,0.5)$, target 3.

```{r}
set.seed(11)
p <- c(0.2, 0.4, 0.6, 0.8, 0.5, 0.5)
r <- replicate(2000, balanced_ra(prob_unit = p))
max(abs(rowMeans(r) - p))
table(colSums(r))
```

Every draw treats exactly 3. The largest unit-level error in 2,000 draws is a few percentage points, which is Monte Carlo noise around an exact marginal.

## Two arms with blocks: overall and within-block

Ten blocks of three, default $p=0.5$. Overall target 15. Each block target 1.5.

```{r}
set.seed(12)
blocks <- rep(1:10, each = 3)
r <- replicate(2000, balanced_ra(blocks = blocks))
table(colSums(r))
block_range <- sapply(1:10, function(b)
  range(colSums(r[blocks == b, , drop = FALSE])))
rownames(block_range) <- c("min", "max")
block_range
max(abs(rowMeans(r) - 0.5))
```

The overall target is the integer 15, and every draw hits 15. Every block contributes 1 or 2.

## Three arms, no blocks: overall tight

The $2 \times 3$ matrix from Section 4, now as a count check. Column targets are $0.62$, $1.13$, and $0.25$, so the tight counts are $0$ or $1$, $1$ or $2$, and $0$ or $1$.

```{r}
set.seed(13)
P <- cbind(c(0.15, 0.47), c(0.65, 0.48), c(0.20, 0.05))
Z <- replicate(2000, as.integer(as.character(
  balanced_ra(prob_unit_each = P, conditions = 1:3))))
count_tab <- sapply(1:3, function(j)
  table(factor(colSums(Z == j), levels = 0:2)))
colnames(count_tab) <- paste0("arm", 1:3)
count_tab
max(abs(rowMeans(Z == 1) - P[, 1]),
    abs(rowMeans(Z == 2) - P[, 2]),
    abs(rowMeans(Z == 3) - P[, 3]))
```

Arm 2 is never empty. No arm moves more than one away from its target. The unit-level shares sit near $P$.

## Two-arm blocked districts, again, as a joint check

```{r}
set.seed(14)
districts <- rep(c("north", "south"), each = 3)
r <- replicate(2000, balanced_ra(blocks = districts))
north <- colSums(r[districts == "north", ])
south <- colSums(r[districts == "south", ])
table(north, south)
all(colSums(r) == 3)
```

The support is $\{(1,2),(2,1)\}$ only. Overall and within-block tightness hold together.

## Equal probabilities and `complete_ra()`

When every unit has the same probability and there are no blocks, `balanced_ra()` is complete assignment. The treated counts match.

```{r}
set.seed(15)
table(colSums(replicate(1000, balanced_ra(4))))
table(colSums(replicate(1000, complete_ra(N = 4))))
```

# 6. Warnings / caveats

`balanced_ra()` is cube/pivotal assignment on the same interface family as `complete_ra()` and `block_ra()`: `N`, `prob_unit`, `blocks`, `clusters`, `conditions`. Exclusive assignment, exact marginals, probabilities in $[0,1]$. For two-arm blocked designs leftover pairing keeps both the within-block counts and the overall count tight. Independently landing each block's leftover had let the overall count wander; that is the two-district case.

The function is experimental. The interface may change. Declare a design with `declare_ra(..., ra_type = "balanced")` or by supplying `prob_unit_each`; `conduct_ra()` and `obtain_condition_probabilities()` then dispatch as they do for the other assignment functions.

Counts are always tight within each block. With two arms they are tight overall as well. With three or more arms the blocks are landed separately, so the overall count can wander. Three blocks of two units, three arms, every probability $1/3$: each tight block gives each arm 0 or 1; the overall target is 2; the three blocks can put an arm at 0 or at 3.

```{r}
set.seed(16)
P <- matrix(1/3, 6, 3)
blocks <- rep(1:3, each = 2)
Z <- replicate(2000, as.integer(as.character(
  balanced_ra(prob_unit_each = P, blocks = blocks, conditions = 1:3))))
table(colSums(Z == 1))
```

Arm 1's overall count is not confined to $\{1,2\}$. That coupling is feasible — Hamilton's method would give each arm to exactly two of the three blocks — and this implementation does not do it when $k>2$.

There is no residual condition if the rows of `prob_unit_each` do not sum to 1. There is no interpretation of a probability greater than 1 as an expected count. Both are refused, as in Section 4.

`check_inputs = FALSE` skips those checks. It is for simulation loops whose probabilities have already been verified. It is not a way to pass illegal designs.

HC2 coverage under these designs is in `vignette("balanced_ra_hc2", package = "randomizr")`.

The default path holds assignment *counts* against their targets. That is not the same as balancing a continuous covariate. Pass `formula` to run cube-on-X on a model matrix; `blocks` tightens counts within discrete groups. The two are different constraints. The function refuses both at once. See `vignette("balanced_ra_covariates", package = "randomizr")`.

The takeaway is practical. Use `balanced_ra()` when probabilities vary across units and the assigned counts should stay as close to their targets as arithmetic allows. Trust overall tightness without blocks, and both overall and within-block tightness with two arms. Do not trust overall tightness with three or more arms and `blocks`.

# References

Deville, J.-C. and Tillé, Y. (1998). Unequal probability sampling without replacement through a splitting method. *Biometrika* 85(1), 89–101. <https://doi.org/10.1093/biomet/85.1.89>

Deville, J.-C. and Tillé, Y. (2004). Efficient balanced sampling: the cube method. *Biometrika* 91(4), 893–912. <https://doi.org/10.1093/biomet/91.4.893>

Deville, J.-C. and Tillé, Y. (2005). Variance approximation under balanced sampling. *Journal of Statistical Planning and Inference* 128(2), 569–591. <https://doi.org/10.1016/j.jspi.2003.11.011>

Chauvet, G. and Tillé, Y. (2006). A fast algorithm for balanced sampling. *Computational Statistics* 21(1), 53–62. <https://doi.org/10.1007/s00180-006-0250-2>
