---
title: "Covariate totals versus blocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariate totals versus blocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
set.seed(20260822)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 3.8
)
options(digits = 3)
```

```{r}
library(randomizr)
library(ggplot2)
```

With `balance_ra` you can balance against a continuous covariate without dividing it into blocks. Cube-on-X with `formula = ~ x` can keeps the covariate imbalance
$T = \sum_i x_i (Z_i - p_i)$ closer to zero than cutting the same $x$ into two,
three, or four blocks. 

We illustrate here with a simple six-unit vector with values
$x = (0, 1, 5, 6, 8, 9)$ and again at $N = 200$ with a continuous covariate.
Blocking keeps the treated count tight inside each bin. The two procedures
impose different linear constraints. Neither dominates for every estimand.

The primary metric is the RMSE of $T$. With heterogeneous $p$ that is the
Horvitz--Thompson residual for the $x$-total; it is also the right comparison
when $p$ is constant. I also report the treated-count residual
$\sum_i Z_i - \sum_i p_i$. `simple_ra()` is a floor: independent assignment,
no balance constraint.

`blocks` remains the partitioned count path in `balanced_ra()`. `formula`
builds a balancing matrix $X$ and flies in $\ker(X')$. Passing both errors.
The count-tight C++ used when `formula` is `NULL` is unchanged.

Section 1 states the two constraints. Section 2 works the six-unit design,
uniform and heterogeneous. Section 3 reports Monte Carlo. The last section
draws the implication.

# 1. Two constraints

The cube method of [Deville and Tillé (2004)](https://doi.org/10.1093/biomet/91.4.893)
holds $X'Z$ near $X'\pi$. The flight phase moves in the kernel of $X'$ until a
coordinate of $Z$ hits 0 or 1. Each step is a martingale, so
$E[Z_i] = p_i$. The landing phase begins when no nonzero kernel move remains
and may drop a constraint. Exact tightness on every column is not always
possible.

`model.matrix(formula, data)` is $X$. `~ x` is an intercept plus $x$. The
intercept *is* the count constraint: $\sum_i Z_i$ stays near $\sum_i p_i$.
The $x$ column asks $\sum_i x_i Z_i$ to stay near $\sum_i x_i p_i$.
`~ 0 + x` drops the intercept; the count may wander.

`blocks` does something else. It tightens treated *counts* inside discrete
groups. Those count constraints only indirectly discipline the mean of $x$.
A factor $B$ belongs in the formula, as `~ x + B`, if the object of interest
is the column space of that model matrix. It does not belong in both
`formula` and `blocks`.

# 2. Six units

```{r}
x <- c(0, 1, 5, 6, 8, 9)
```

Three groupings follow the gaps in $x$. The three-block cut keeps the natural
pairs. The two-block cut has to split the middle pair. The four-block cut
splits that pair into singletons.

```{r}
br2 <- c(-Inf, 5.5, Inf)
br3 <- c(-Inf, 3, 7, Inf)
br4 <- c(-Inf, 3, 5.5, 7, Inf)
b2 <- cut(x, br2, labels = c("(0,1,5)", "(6,8,9)"))
b3 <- cut(x, br3, labels = c("(0,1)", "(5,6)", "(8,9)"))
b4 <- cut(x, br4, labels = c("(0,1)", "(5)", "(6)", "(8,9)"))
data.frame(x, b2, b3, b4)
```

## Uniform probabilities

With $p_i = 0.5$ the target treated count is 3. The target $x$-total is
$0.5 \times \sum x = 14.5$. That value is not an integer, so no 0-1 assignment
hits it. Landing must sit on a nearby vertex.

Three-block complete assignment treats exactly one unit in each pair. The
eight assignments and their $x$-totals:

```{r}
pairs <- list(c(0, 1), c(5, 6), c(8, 9))
grid <- expand.grid(pairs)
grid$sum_x <- rowSums(grid)
grid
```

So three-block RA produces $\sum x Z \in \{13, 14, 15, 16\}$. Cube-on-X
targets 14.5 and, after landing, concentrates on 14 and 15.

```{r}
p_u <- rep(0.5, 6)
S <- 2000
draw_form <- function(p) balanced_ra(prob_unit = p, formula = ~ x)
draw_b <- function(p, b) balanced_ra(prob_unit = p, blocks = b)
z_form_u <- replicate(S, draw_form(p_u))
z_b2_u   <- replicate(S, draw_b(p_u, b2))
z_b3_u   <- replicate(S, draw_b(p_u, b3))
z_b4_u   <- replicate(S, draw_b(p_u, b4))
z_cnt_u  <- replicate(S, balanced_ra(prob_unit = p_u))
z_sim_u  <- replicate(S, simple_ra(N = 6, prob_unit = p_u))

T_of <- function(Z, x, p) colSums(x * (Z - p))
C_of <- function(Z, p) colSums(Z) - sum(p)

rmse <- function(v) sqrt(mean(v^2))
mae  <- function(v) mean(abs(v))

metrics <- function(Z, x, p) {
  T <- T_of(Z, x, p)
  C <- C_of(Z, p)
  c(rmse_T = rmse(T), mae_T = mae(T), rmse_C = rmse(C), mae_C = mae(C))
}

round(rbind(
  formula   = metrics(z_form_u, x, p_u),
  blocks_2  = metrics(z_b2_u, x, p_u),
  blocks_3  = metrics(z_b3_u, x, p_u),
  blocks_4  = metrics(z_b4_u, x, p_u),
  counts    = metrics(z_cnt_u, x, p_u),
  simple    = metrics(z_sim_u, x, p_u)
), 3)
```

Cube-on-X has the smallest RMSE of $T$. Unblocked `balanced_ra()` matches it
on the count residual (the intercept) and loses on $T$. Three-block RA is
tighter on within-pair counts and worse on $T$, because it admits 13 and 16.

```{r}
support <- function(Z, x) {
  s <- apply(Z, 2, function(z) paste(z, collapse = ""))
  tot <- colSums(x * Z)
  tab <- sort(tapply(tot, s, function(v) v[1]))
  data.frame(assignment = names(tab), sum_xZ = as.numeric(tab),
             n = as.integer(table(s)[names(tab)]), row.names = NULL)
}
cat("formula = ~ x\n")
support(z_form_u, x)
cat("3-block\n")
support(z_b3_u, x)
```

The three-block support is the eight pair-wise assignments. Cube-on-X puts
almost all mass on count-3 assignments whose $x$-total is 14 or 15. Those
two sets overlap; they are not the same.

## Heterogeneous probabilities

```{r}
p_h <- c(0.2, 0.3, 0.4, 0.6, 0.7, 0.8)
sum(p_h)
sum(x * p_h)
```

The target count is still 3. The target $x$-total is 18.7, not 14.5. Blocks
with heterogeneous $p$ inside a bin are a different object from putting $x$
in $X$. In the middle pair the unit probabilities are 0.4 and 0.6, so
three-block RA treats exactly one of $\{5, 6\}$ and ignores that 6 is larger
and more likely.

```{r}
z_form_h <- replicate(S, draw_form(p_h))
z_b2_h   <- replicate(S, draw_b(p_h, b2))
z_b3_h   <- replicate(S, draw_b(p_h, b3))
z_b4_h   <- replicate(S, draw_b(p_h, b4))
z_cnt_h  <- replicate(S, balanced_ra(prob_unit = p_h))
z_sim_h  <- replicate(S, simple_ra(N = 6, prob_unit = p_h))

round(rbind(
  formula   = metrics(z_form_h, x, p_h),
  blocks_2  = metrics(z_b2_h, x, p_h),
  blocks_3  = metrics(z_b3_h, x, p_h),
  blocks_4  = metrics(z_b4_h, x, p_h),
  counts    = metrics(z_cnt_h, x, p_h),
  simple    = metrics(z_sim_h, x, p_h)
), 3)
```

The ranking on $T$ is the same. Cube-on-X tracks 18.7. Three-block RA still
forces one treated unit in $(5, 6)$, which is the wrong linear constraint
for that target.

```{r, fig.cap="Imbalance T under uniform p = 0.5 (left) and heterogeneous p (right). Cube-on-X sits closer to zero than the blocked designs."}
plot_T <- function(designs, title) {
  d <- do.call(rbind, lapply(names(designs), function(nm) {
    data.frame(design = nm, T = designs[[nm]])
  }))
  d$design <- factor(d$design, names(designs))
  ggplot(d, aes(T, fill = design)) +
    geom_histogram(binwidth = 1, position = "identity", alpha = 0.45,
                   color = "white") +
    geom_vline(xintercept = 0, linetype = 2, linewidth = 0.3) +
    scale_fill_brewer(palette = "Dark2") +
    labs(x = "T = sum x (Z - p)", y = "Draws", title = title) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom", legend.title = element_blank())
}
p_left <- plot_T(list(
  "formula ~ x" = T_of(z_form_u, x, p_u),
  "3 blocks"    = T_of(z_b3_u, x, p_u),
  "counts only" = T_of(z_cnt_u, x, p_u)
), "Uniform p = 0.5")
p_right <- plot_T(list(
  "formula ~ x" = T_of(z_form_h, x, p_h),
  "3 blocks"    = T_of(z_b3_h, x, p_h),
  "counts only" = T_of(z_cnt_h, x, p_h)
), "Heterogeneous p")
p_left
p_right
```

# 3. A larger design

The same comparison at $N = 200$, with $x$ continuous. Quantile bins play the
role that `ntile(x, 2/3/4)` plays in a blocked design. $S = 500$ draws.

```{r}
N <- 200
S2 <- 500
x200 <- rnorm(N)
p200 <- plogis(-0.3 + 0.4 * x200)
ntile <- function(v, n) {
  rnk <- rank(v, ties.method = "first")
  as.integer(floor((rnk - 1) / length(v) * n) + 1L)
}

z_f  <- replicate(S2, balanced_ra(prob_unit = p200, formula = ~ x200))
z_2  <- replicate(S2, balanced_ra(prob_unit = p200, blocks = ntile(x200, 2)))
z_3  <- replicate(S2, balanced_ra(prob_unit = p200, blocks = ntile(x200, 3)))
z_4  <- replicate(S2, balanced_ra(prob_unit = p200, blocks = ntile(x200, 4)))
z_c  <- replicate(S2, balanced_ra(prob_unit = p200))
z_s  <- replicate(S2, simple_ra(N = N, prob_unit = p200))

lab <- c("formula ~ x", "2 blocks", "3 blocks", "4 blocks",
         "counts only", "simple")
M <- rbind(
  metrics(z_f, x200, p200),
  metrics(z_2, x200, p200),
  metrics(z_3, x200, p200),
  metrics(z_4, x200, p200),
  metrics(z_c, x200, p200),
  metrics(z_s, x200, p200)
)
rownames(M) <- lab
round(M, 3)
```

```{r, fig.cap="RMSE of T at N = 200. Formula wins on the linear-in-x total. Blocking wins on within-bin counts (not shown); more bins helps T only insofar as the bins approximate x."}
bar <- data.frame(design = factor(lab, lab), rmse_T = M[, "rmse_T"])
ggplot(bar, aes(design, rmse_T)) +
  geom_col(width = 0.65, fill = "#0054AD") +
  labs(x = NULL, y = "RMSE of T") +
  theme_bw(base_size = 11)
```

More bins shrink $T$ relative to two blocks, because the piecewise-constant
count constraints become a coarser stand-in for $x$. They do not catch the
formula, which puts $x$ in $X$ directly. Count-only `balanced_ra()` and
`simple_ra()` remain the worst on $T$; they were not asked to balance $x$.

Within-bin counts go the other way. On the three-block design every draw
keeps each tercile at the floor or ceiling of its target. Cube-on-X does not
promise that. The lesson is the constraint, not a ranking that holds for
every estimand.

# 4. What to take away

Use `formula = ~ x` when the object you care about is a linear-in-$x$ total.
Use `blocks` when you care about treated counts inside discrete groups.
Do not pass both. Multi-arm `prob_unit_each` with `formula` is not
implemented.

# References

Deville, J.-C. and Tillé, Y. (2004). Efficient balanced sampling: the cube
method. *Biometrika* 91(4), 893–912. <https://doi.org/10.1093/biomet/91.4.893>

Chauvet, G. and Tillé, Y. (2006). A fast algorithm for balanced sampling.
*Computational Statistics* 21(1), 53–62.
<https://doi.org/10.1007/s00180-006-0250-2>
