---
title: "HC2 coverage under balanced assignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HC2 coverage under balanced assignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
set.seed(20260822)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.4
)
options(digits = 3)
```

```{r}
library(randomizr)
library(estimatr)
library(ggplot2)
```

Under the designs simulated here, HC2 95 percent intervals cover the ATE at 0.945 (two-arm), 0.941 (blocked two-arm), and 0.954 (three-arm), each from 2,000 draws. Independent Bernoulli assignment on the same two-arm DGP covers at 0.935. Assignment dependence is real: the mean pairwise correlation is $-0.20$ under balanced assignment and $0.00$ under simple assignment, and the treated count does not wander. That dependence does not throw HC2 off on these designs. Two-arm coverage sits a shade below 95 percent; three-arm coverage sits a shade above. Section 1 states why one might have worried. Section 2 reports the simulations. The claim is limited to the designs below. It is not a theorem about every cube design.

# 1. Why one might have worried

`simple_ra()` draws each unit independently. Pairwise assignment correlations are zero. The treated count wanders.

`complete_ra()` and `block_ra()` fix the treated count, or the count within each block. Pairwise assignment is then negatively dependent. The usual two-arm complete design has $\mathrm{Corr}(Z_i,Z_j)=-1/(N-1)$. That is still a design in which every unit has the same probability.

`balanced_ra()` keeps unit-varying probabilities *and* holds the counts tight. The cube walk (two-arm: the pivotal method) pairs leftover units. With `blocks`, two-arm leftover pairing also couples units across blocks. Pairwise assignment is not independent Bernoulli. That is the point of the design: less assignment noise. It is also why people ask whether heteroskedasticity-consistent intervals, built from a working independence model, still cover.

The estimator here is inverse-probability-weighted least squares with HC2 standard errors ([MacKinnon and White 1985](https://doi.org/10.1016/0304-4076(85)90158-7)), via `estimatr::lm_robust(..., se_type = "HC2")`. When $p_i$ varies, unweighted difference-in-means is not the ATE. The weights are $Z_i/p_i + (1-Z_i)/(1-p_i)$.

The next block makes the dependence visible on a six-unit design. `balanced_ra()` holds the treated count at 3. `simple_ra()` does not. Off-diagonal assignment correlations are near zero under simple assignment and about $-0.20$ under balanced assignment.

```{r}
p <- c(0.2, 0.4, 0.6, 0.8, 0.5, 0.5)
n_dep <- 4000L
Z_bal <- replicate(n_dep, balanced_ra(prob_unit = p, check_inputs = FALSE))
Z_sim <- replicate(n_dep, simple_ra(N = 6, prob_unit = p,
                                    conditions = 0:1, check_inputs = FALSE))

c(balanced_treated = var(colSums(Z_bal)),
  simple_treated   = var(colSums(Z_sim)))

mean_offdiag <- function(S) {
  C <- cor(t(S))
  mean(C[upper.tri(C)])
}
c(balanced_pair_cor = mean_offdiag(Z_bal),
  simple_pair_cor   = mean_offdiag(Z_sim))
```

# 2. Coverage

Potential outcomes are fixed. Treatment effects are constant. $Y_i(0)$ is correlated with $p_i$, so inverse-probability weights are needed. Each cell redraws assignment from the current `balanced_ra()` implementation, estimates the ATE by IPW least squares, and records whether the HC2 interval covers the true ATE. Tight loops set `check_inputs = FALSE`. Two thousand draws for the three main designs; one thousand for the grids. Monte Carlo error on a 95 percent coverage rate at 2,000 draws is about half a percentage point.

```{r}
cover_ipw <- function(assign, p, y0, tau, nrep) {
  cover <- logical(nrep)
  se <- est <- numeric(nrep)
  for (r in seq_len(nrep)) {
    Z <- assign()
    Y <- y0 + tau * Z
    w <- Z / p + (1 - Z) / (1 - p)
    fit <- lm_robust(Y ~ Z, weights = w, se_type = "HC2")
    est[r] <- fit$coefficients[["Z"]]
    se[r]  <- fit$std.error[["Z"]]
    cover[r] <- fit$conf.low[["Z"]] <= tau && tau <= fit$conf.high[["Z"]]
  }
  data.frame(
    coverage = mean(cover),
    mean_est = mean(est),
    mean_se  = mean(se),
    nrep = nrep,
    stringsAsFactors = FALSE
  )
}

tau <- 1
nrep_main <- 2000L
```

## Two-arm, unit-varying $p$

$N=80$. $p_i$ runs from 0.2 to 0.8. $Y_i(0)=2p_i+\varepsilon_i$. The same potential outcomes are also assigned with `simple_ra()`, so the comparison isolates assignment dependence.

```{r}
N <- 80L
p80 <- seq(0.2, 0.8, length.out = N)
y0_80 <- 2 * p80 + rnorm(N)

two_arm <- rbind(
  cbind(design = "balanced",
        cover_ipw(function() balanced_ra(prob_unit = p80, check_inputs = FALSE),
                  p80, y0_80, tau, nrep_main)),
  cbind(design = "simple",
        cover_ipw(function() simple_ra(N = N, prob_unit = p80,
                                      conditions = 0:1, check_inputs = FALSE),
                  p80, y0_80, tau, nrep_main))
)
two_arm
```

## Blocked two-arm

Eight blocks of ten. $p_i$ varies within each block. Counts are tight within blocks and overall.

```{r}
blocks <- rep(seq_len(8L), each = 10L)
p_blk <- as.vector(replicate(8L, seq(0.25, 0.75, length.out = 10L)))
y0_blk <- 2 * p_blk + rnorm(length(p_blk))

blocked <- cbind(
  design = "balanced, blocked",
  cover_ipw(function() balanced_ra(prob_unit = p_blk, blocks = blocks,
                               check_inputs = FALSE),
            p_blk, y0_blk, tau, nrep_main)
)
blocked
```

## Three-arm

$N=90$. Each unit has its own row of probabilities. The estimand is the T2-versus-T1 contrast. Weights are the reciprocal of the probability of the condition the unit landed in.

```{r}
cover_multi <- function(P, y0, tau, nrep) {
  cover <- est <- se <- numeric(nrep)
  cond <- colnames(P) <- c("T1", "T2", "T3")
  for (r in seq_len(nrep)) {
    Z <- balanced_ra(prob_unit_each = P, conditions = cond, check_inputs = FALSE)
    Y <- y0 + tau * as.integer(Z == "T2")
    w <- P[cbind(seq_len(nrow(P)), match(as.character(Z), cond))]
    fit <- lm_robust(Y ~ Z, weights = 1 / w, se_type = "HC2")
    est[r] <- fit$coefficients[["ZT2"]]
    se[r]  <- fit$std.error[["ZT2"]]
    cover[r] <- fit$conf.low[["ZT2"]] <= tau && tau <= fit$conf.high[["ZT2"]]
  }
  data.frame(
    coverage = mean(cover),
    mean_est = mean(est),
    mean_se  = mean(se),
    nrep = nrep,
    stringsAsFactors = FALSE
  )
}

N3 <- 90L
u <- seq(0, 1, length.out = N3)
P3 <- cbind(T1 = 0.20 + 0.20 * u,
            T2 = 0.45 - 0.10 * u,
            T3 = 0.35 - 0.10 * u)
P3 <- P3 / rowSums(P3)
y0_3 <- 2 * P3[, "T2"] + rnorm(N3)

multi <- cbind(design = "balanced, three-arm",
               cover_multi(P3, y0_3, tau, nrep_main))
multi
```

```{r}
results <- rbind(two_arm, blocked, multi)
results
```

## Coverage against $N$ and against the spread of $p$

The same two-arm DGP, four sample sizes, 1,000 draws each. Then $N=80$ with four spreads of $p_i$ around 0.5. Neither grid moves coverage far from 95 percent.

```{r}
nrep_grid <- 1000L
Ns <- c(20L, 40L, 80L, 160L)
by_N <- do.call(rbind, lapply(Ns, function(n) {
  p <- seq(0.25, 0.75, length.out = n)
  y0 <- 2 * p + rnorm(n)
  cbind(N = n,
        cover_ipw(function() balanced_ra(prob_unit = p, check_inputs = FALSE),
                  p, y0, tau, nrep_grid))
}))
by_N

spreads <- c(0, 0.15, 0.30, 0.40)
by_spread <- do.call(rbind, lapply(spreads, function(a) {
  p <- pmin(0.9, pmax(0.1, 0.5 + a * seq(-1, 1, length.out = 80L)))
  y0 <- 2 * p + rnorm(80L)
  cbind(spread = a, sd_p = sd(p),
        cover_ipw(function() balanced_ra(prob_unit = p, check_inputs = FALSE),
                  p, y0, tau, nrep_grid))
}))
by_spread
```

```{r, fig.cap = "HC2 coverage of the ATE against $N$ and against the spread of $p_i$. Dashed line: nominal 95 percent."}
grid <- rbind(
  data.frame(panel = "N", x = by_N$N, coverage = by_N$coverage),
  data.frame(panel = "spread of p", x = by_spread$spread, coverage = by_spread$coverage)
)
ggplot(grid, aes(x, coverage)) +
  geom_hline(yintercept = 0.95, linetype = 2) +
  geom_point() +
  geom_line() +
  facet_wrap(~ panel, scales = "free_x") +
  coord_cartesian(ylim = c(0.90, 1)) +
  labs(x = NULL, y = "HC2 coverage")
```

# 3. What this does not say

These are the designs that were drawn: two-arm unit-varying $p$, the same design assigned independently, blocked two-arm with $p$ varying inside blocks, and a three-arm cube design. HC2 intervals sat near 95 percent in each case. Two-arm rates were 0.94–0.95; the three-arm rate was 0.954. Assignment dependence is visible and does not wreck the intervals here.

The note does not say that HC2 is valid for every cube design. It does not speak to cluster-robust standard errors, to randomization inference, or to designs that empty an arm. Use the simulation if a design sits far from these.

The takeaway is narrow. Tight counts induce dependence. On the designs above, that is not a reason to drop HC2.
