---
title: "Covariate totals versus blocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariate totals versus blocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# pak::pak("DeclareDesign/fabricatr@rewrite")
# pak::pak("DeclareDesign/DeclareDesign@rewrite")
# pak::pak("macartan/ResearchDesigns@coppock/param-plumbing")
```

```{r, include = FALSE}
set.seed(20260822)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 3.8
)
options(digits = 3)
```

```{r}
library(fabricatrZero)
library(DeclareDesignZero)
library(randomizr)
library(ggplot2)
library(estimatr)
library(rlang)
```

Rather than using blocks we can assign against a continuous covariate.

This vignette compares an approach in which a continuous $X$ is used directly and an approach in which $X$ is split into blocks.

On the six-unit vector $x = (0, 1, 5, 6, 8, 9)$ the three-block cut is already a tight proxy for $x$: one treated unit in each pair puts $\sum x Z$ in $\{13, 14, 15, 16\}$. Cube-on-X targets $14.5$ and beats `complete_ra`; it does not beat that particular blocking on $T$. With heterogeneous $p$ the target is $18.7$, and cube-on-X and blocking part company.

When every unit has the same $p$, `complete_ra` and `balanced_ra()` with no formula are the same design: a fixed treated count, no use of $x$. When $p$ varies they are not. `complete_ra` uses a scalar $m$ (or `prob`) and ignores unit-level $p$. `balanced_ra()` with no formula honors $p_i$ and still tightens $\sum Z$. Neither uses $x$. `simple_ra()` is independent Bernoulli assignment: no count constraint and no use of $x$.

For a larger case with $N = 200$ and a continuous covariate, `balanced_ra(formula = ~ x)` keeps the imbalance $T = \sum_i x_i (Z_i - p_i)$ smaller than cutting $x$ into two, three, or four quantile blocks. Blocking, however, keeps the treated count tight inside each bin. Those are different constraints. Neither dominates for every estimand.

The primary metric is the RMSE of $T$. With heterogeneous $p$ that is the Horvitz--Thompson residual for the $x$-total; it is also the right comparison when $p$ is constant. We also report the treated-count residual $\sum_i Z_i - \sum_i p_i$.

`blocks` remains the partitioned count path in `balanced_ra()`. `formula` builds a balancing matrix $X$ and flies in $\ker(X')$. Passing both errors. The count-tight C++ used when `formula` is `NULL` is unchanged.

When $Y = x^2$, so that $x$ is strongly prognostic, most of the RMSE gain is from controlling for $x$, or for bins of $x$, rather than from the assignment. Cube-on-X plus Lin still has the smallest RMSE, and the assignment is what separates it from the same estimator after complete assignment. Unadjusted DiM after cube-on-X does not share that gain.

Section 1 states the two constraints and situates cube-on-X among related designs. Section 2 works the six-unit design, uniform and heterogeneous. Section 3 reports Monte Carlo. Section 4 diagnoses six assignment--estimator pairs when $Y$ is quadratic in $x$. The last section draws the implication.

# 1. Two constraints

The cube method of [Deville and Tillé (2004)](https://doi.org/10.1093/biomet/91.4.893) holds $X'Z$ near $X'\pi$. The flight phase moves in the kernel of $X'$ until a coordinate of $Z$ hits 0 or 1. Each step is a martingale, so $E[Z_i] = p_i$. The landing phase begins when no nonzero kernel move remains and may drop a constraint. Exact tightness on every column is not always possible.

`model.matrix(formula, data)` is $X$. `~ x` is an intercept plus $x$. The intercept *is* the count constraint: $\sum_i Z_i$ stays near $\sum_i p_i$. The $x$ column asks $\sum_i x_i Z_i$ to stay near $\sum_i x_i p_i$. `~ 0 + x` drops the intercept; the count may wander.

`blocks` does something else. It tightens treated *counts* inside discrete groups. Those count constraints only indirectly constrain the mean of $x$. A factor $B$ belongs in the formula, as `~ x + B`, if the object of interest is the column space of that model matrix. It does not belong in both `formula` and `blocks`.

## Related designs

Cube-on-X is one way to balance a continuous covariate without cutting it into bins. It is not the only way, and the two-arm cube already lives in the survey stack.

[Deville and Tillé (2004)](https://doi.org/10.1093/biomet/91.4.893) is implemented for sampling as `sampling::samplecube` and `BalancedSampling::cube`. [Davezies, Hollard, and Vergara Merino (2024)](https://arxiv.org/abs/2407.13613) use `BalancedSampling` for RCTs; they do not ship a separate package.

[Harshaw, Sävje, Spielman, and Zhang (2024)](https://doi.org/10.1080/01621459.2023.2285474) give a different two-arm design, the Gram--Schmidt walk. It balances continuous $X$ and trades robustness against covariate balance through a parameter $\phi$. It is not the cube, and unit-varying $\pi$ is not a first-class input. Software is the Julia package `GSWDesign.jl` and the R wrapper `gswdesign` (not on CRAN).

Rerandomization ([Morgan and Rubin 2012](https://doi.org/10.1214/12-AOS1008)) rejects complete RA until $X$ is balanced; see, for example, `fastrerandomize`. Match-then-randomize and threshold blocking (`blockTools`, `quickblock`; [Higgins, Sävje, and Sekhon 2016](https://doi.org/10.1073/pnas.1510504113)) form bins or pairs and then complete RA inside. `minMSE` ([Schneider and Schlather 2017](https://www.econstor.eu/handle/10419/161931)) optimizes the assignment; that is a different algorithm.

`balanced_ra(formula = ~ x)` is the cube with a randomizr interface: leftover pairing, `blocks`, and the rest of the multi-arm tools. It is not a new sampling method.

# 2. Six units

```{r}
x <- c(0, 1, 5, 6, 8, 9)
```

Three groupings follow the gaps in $x$. The three-block cut keeps the natural pairs. The two-block cut has to split the middle pair. The four-block cut splits that pair into singletons.

```{r}
b2 <- cut(x, c(-Inf, 5.5, Inf), labels = c("(0,1,5)", "(6,8,9)"))
b3 <- cut(x, c(-Inf, 3, 7, Inf), labels = c("(0,1)", "(5,6)", "(8,9)"))
b4 <- cut(x, c(-Inf, 3, 5.5, 7, Inf), labels = c("(0,1)", "(5)", "(6)", "(8,9)"))
knitr::kable(
  data.frame(x, `2 blocks` = b2, `3 blocks` = b3, `4 blocks` = b4,
             check.names = FALSE),
  caption = "Cuts of the six-unit $x$."
)
```

## Uniform probabilities

With $p_i = 0.5$ the target treated count is 3. The target $x$-total is $0.5 \times \sum x = 14.5$. That value is not an integer, so no 0-1 assignment hits it. Landing must sit on a nearby vertex.

Three-block complete assignment treats exactly one unit in each pair. The eight assignments and their $x$-totals:

```{r}
pairs <- list(c(0, 1), c(5, 6), c(8, 9))
grid <- expand.grid(pairs)
names(grid) <- c("pair 1", "pair 2", "pair 3")
grid$sum_xZ <- rowSums(grid)
knitr::kable(grid, caption = "Treated $x$ values and $\\sum x Z$ under three-block assignment.")
```

Three-block RA therefore produces $\sum x Z \in \{13, 14, 15, 16\}$. Cube-on-X targets 14.5. Landing may drop the $x$ column, so a draw need not sit on 14 or 15. The nearest feasible 0-1 totals are still 14 and 15; they just are not the whole support.

With constant $p = 0.5$, `balanced_ra()` with no formula is the same design as `complete_ra`: it tightens $\sum Z$ and ignores $x$. The baseline below is therefore `complete_ra`. `simple_ra` assigns each unit independently at $p = 0.5$.

```{r}
p_u <- rep(0.5, 6)
S <- 2000
draw_form <- function(p) balanced_ra(prob_unit = p, formula = ~ x)
draw_b <- function(p, b) balanced_ra(prob_unit = p, blocks = b)
z_form_u <- replicate(S, draw_form(p_u))
z_b2_u   <- replicate(S, draw_b(p_u, b2))
z_b3_u   <- replicate(S, draw_b(p_u, b3))
z_b4_u   <- replicate(S, draw_b(p_u, b4))
z_comp_u <- replicate(S, complete_ra(N = 6, m = 3))
z_sim_u  <- replicate(S, simple_ra(N = 6, prob_unit = p_u))

T_of <- function(Z, x, p) colSums(x * (Z - p))
C_of <- function(Z, p) colSums(Z) - sum(p)

rmse <- function(v) sqrt(mean(v^2))
mae  <- function(v) mean(abs(v))

metrics <- function(Z, x, p) {
  T <- T_of(Z, x, p)
  C <- C_of(Z, p)
  c(rmse_T = rmse(T), mae_T = mae(T), rmse_C = rmse(C), mae_C = mae(C))
}

metrics_df <- function(designs, x, p) {
  M <- do.call(rbind, lapply(designs, function(Z) metrics(Z, x, p)))
  data.frame(
    Strategy = names(designs),
    `RMSE of T` = M[, "rmse_T"],
    `MAE of T` = M[, "mae_T"],
    `RMSE of count residual` = M[, "rmse_C"],
    `MAE of count residual` = M[, "mae_C"],
    row.names = NULL,
    check.names = FALSE
  )
}

cap_T <- paste(
  "RMSE and MAE of $T = \\sum_i x_i (Z_i - p_i)$",
  "and of the treated-count residual $\\sum_i Z_i - \\sum_i p_i$."
)

designs_u <- list(
  "formula ~ x" = z_form_u,
  "2 blocks"    = z_b2_u,
  "3 blocks"    = z_b3_u,
  "4 blocks"    = z_b4_u,
  "complete_ra" = z_comp_u,
  "simple_ra"   = z_sim_u
)
knitr::kable(metrics_df(designs_u, x, p_u), digits = 3, align = "lrrrr",
             caption = cap_T)
```

Three- and four-block RA have the smallest RMSE of $T$ on this vector: the bins are close pairs (or singletons), so a within-bin count constraint is already a tight stand-in for $x$. Cube-on-X matches the count residual (the intercept is in $X$) and beats `complete_ra` and `simple_ra` on $T$. It does not beat those cuts. That ranking is special to this $x$, not a general rule. The $N = 200$ design below reverses it.

```{r}
support <- function(Z, x) {
  s <- apply(Z, 2, function(z) paste(z, collapse = ""))
  tot <- colSums(x * Z)
  tab <- sort(tapply(tot, s, function(v) v[1]))
  data.frame(assignment = names(tab), sum_xZ = as.numeric(tab),
             n = as.integer(table(s)[names(tab)]), row.names = NULL)
}
knitr::kable(support(z_form_u, x),
             caption = "Support of `balanced_ra(formula = ~ x)` under uniform $p = 0.5$.")
knitr::kable(support(z_b3_u, x),
             caption = "Support of three-block assignment under uniform $p = 0.5$.")
```

The three-block support is the eight pair-wise assignments. Cube-on-X puts more mass on 14 and 15 than `complete_ra` does, and it also lands on other count-3 totals when the last leftover pair is far apart in $x$. The two supports overlap; they are not the same.

## Heterogeneous probabilities

```{r}
p_h <- c(0.2, 0.3, 0.4, 0.6, 0.7, 0.8)
sum(p_h)
sum(x * p_h)
```

The target count is still 3. The target $x$-total is 18.7, not 14.5. Blocks with heterogeneous $p$ inside a bin are a different object from putting $x$ in $X$. In the middle pair the unit probabilities are 0.4 and 0.6, so three-block RA treats exactly one of $\{5, 6\}$ and ignores that 6 is larger and more likely.

`complete_ra` is no longer the same as count-tight `balanced_ra`. `complete_ra` fixes the treated count at a scalar $m$ (here $m = 3$). It ignores $x$ and ignores unit-level $p$. `balanced_ra` with no formula (counts only) honors $p_i$ and still tightens $\sum Z$. It also ignores $x$. `simple_ra` assigns independently at $p_i$.

```{r}
z_form_h <- replicate(S, draw_form(p_h))
z_b2_h   <- replicate(S, draw_b(p_h, b2))
z_b3_h   <- replicate(S, draw_b(p_h, b3))
z_b4_h   <- replicate(S, draw_b(p_h, b4))
z_comp_h <- replicate(S, complete_ra(N = 6, m = 3))
z_cnt_h  <- replicate(S, balanced_ra(prob_unit = p_h))
z_sim_h  <- replicate(S, simple_ra(N = 6, prob_unit = p_h))

designs_h <- list(
  "formula ~ x"          = z_form_h,
  "2 blocks"             = z_b2_h,
  "3 blocks"             = z_b3_h,
  "4 blocks"             = z_b4_h,
  "complete_ra"          = z_comp_h,
  "balanced_ra (counts)" = z_cnt_h,
  "simple_ra"            = z_sim_h
)
knitr::kable(metrics_df(designs_h, x, p_h), digits = 3, align = "lrrrr",
             caption = cap_T)
```

Cube-on-X tracks 18.7. Three-block RA still forces one treated unit in $(5, 6)$, which is the wrong linear constraint for that target. The RMSE gap is smaller than at $N = 200$; the toy is small enough that landing noise is a large share of $T$. `complete_ra` and count-tight `balanced_ra` both hold $\sum Z$ near 3. They differ on $T$ because only the latter respects $p_i$.

```{r}
metric_long <- function(designs, x, p) {
  out <- do.call(rbind, lapply(names(designs), function(nm) {
    Z <- designs[[nm]]
    rbind(
      data.frame(strategy = nm, metric = "T",
                 value = as.numeric(T_of(Z, x, p))),
      data.frame(strategy = nm, metric = "Count residual",
                 value = as.numeric(C_of(Z, p)))
    )
  }))
  out$strategy <- factor(out$strategy, names(designs))
  out$metric <- factor(out$metric, c("T", "Count residual"))
  out
}

plot_metrics <- function(designs, x, p) {
  d <- metric_long(designs, x, p)
  ggplot(d, aes(value)) +
    geom_histogram(binwidth = 1, fill = "grey55", color = "white",
                   boundary = 0) +
    geom_vline(xintercept = 0, linetype = 2, linewidth = 0.3) +
    facet_grid(strategy ~ metric, scales = "free_x") +
    labs(x = NULL, y = "Draws") +
    theme_bw(base_size = 11) +
    theme(strip.text.y.right = element_text(angle = 0, hjust = 0))
}
```

```{r, fig.height=6.2, fig.cap="Imbalance $T = \\sum_i x_i (Z_i - p_i)$ and the treated-count residual under uniform $p = 0.5$. One row per strategy; $x$-scale shared down each column. On this vector three-block RA is a tight proxy for $x$. Cube-on-X still beats `complete_ra`."}
plot_metrics(designs_u[c("formula ~ x", "3 blocks", "complete_ra", "simple_ra")],
             x, p_u)
```

```{r, fig.height=7.4, fig.cap="Same metrics under heterogeneous $p$. `complete_ra` fixes the count at a scalar $m$ and ignores unit-level $p$. `balanced_ra` (counts) honors $p_i$ and tightens $\\sum Z$; it still ignores $x$."}
plot_metrics(designs_h[c("formula ~ x", "3 blocks", "complete_ra",
                         "balanced_ra (counts)", "simple_ra")],
             x, p_h)
```

# 3. A larger design

The same comparison at $N = 200$, with $x$ continuous. Quantile bins play the role that `ntile(x, 2/3/4)` plays in a blocked design. $S = 500$ draws. Probabilities vary with $x$, so the panel includes both `complete_ra` (scalar `prob`) and count-tight `balanced_ra`.

```{r}
N <- 200
S2 <- 500
x200 <- rnorm(N)
p200 <- plogis(-0.3 + 0.4 * x200)
ntile <- function(v, n) {
  rnk <- rank(v, ties.method = "first")
  as.integer(floor((rnk - 1) / length(v) * n) + 1L)
}

z_f    <- replicate(S2, balanced_ra(prob_unit = p200, formula = ~ x200))
z_2    <- replicate(S2, balanced_ra(prob_unit = p200, blocks = ntile(x200, 2)))
z_3    <- replicate(S2, balanced_ra(prob_unit = p200, blocks = ntile(x200, 3)))
z_4    <- replicate(S2, balanced_ra(prob_unit = p200, blocks = ntile(x200, 4)))
z_comp <- replicate(S2, complete_ra(N = N, prob = mean(p200)))
z_c    <- replicate(S2, balanced_ra(prob_unit = p200))
z_s    <- replicate(S2, simple_ra(N = N, prob_unit = p200))

designs_n <- list(
  "formula ~ x"          = z_f,
  "2 blocks"             = z_2,
  "3 blocks"             = z_3,
  "4 blocks"             = z_4,
  "complete_ra"          = z_comp,
  "balanced_ra (counts)" = z_c,
  "simple_ra"            = z_s
)
knitr::kable(metrics_df(designs_n, x200, p200), digits = 3, align = "lrrrr",
             caption = cap_T)
```

```{r, fig.height=8.6, fig.cap="Distributions of $T = \\sum_i x_i (Z_i - p_i)$ and of the treated-count residual at $N = 200$. Formula wins on the linear-in-$x$ total. Blocking wins on within-bin counts (not shown); more bins help $T$ only insofar as the bins approximate $x$."}
plot_metrics(designs_n, x200, p200)
```

More bins shrink $T$ relative to two blocks, because the piecewise-constant count constraints become a coarser stand-in for $x$. They do not catch the formula, which puts $x$ in $X$ directly. `complete_ra`, count-tight `balanced_ra`, and `simple_ra` remain the worst on $T$; they were not asked to balance $x$. `complete_ra` is worse than count-tight `balanced_ra` here because $p$ rises with $x$ and complete assignment ignores that.

Within-bin counts go the other way. On the three-block design every draw keeps each tercile at the floor or ceiling of its target. Cube-on-X does not promise that. The lesson is the constraint, not a ranking that holds for every estimand.

# 4. Estimator efficiency when $Y$ is quadratic in $x$

The comparisons above are about $T = \sum_i x_i (Z_i - p_i)$. They do not say which assignment--estimator pair is most efficient for the ATE.

$x$ is strongly prognostic: $Y(0) = x^2$. Most of the RMSE gain in this section comes from controlling for $x$, or for bins of $x$, rather than from the assignment. What the design adds sits on top of that. Cube-on-X plus Lin has the smallest RMSE. Complete assignment plus unadjusted DiM has the largest.

*Analyze as you randomize* remains the default pairing: block dummies with blocked assignment, $x$ in the formula with Lin. Because $x$ is so strongly prognostic, the section also reports misspecified and extra controls: block dummies after complete assignment, linear $x$ after blocking, and unadjusted DiM after cube-on-X.

This section diagnoses a two-arm trial with $N = 10$, $x = 1, \ldots, 10$, $Y(0) = x^2$, $Y(1) = x^2 + 10$. The ATE is 10. Six assignment--estimator pairs:

1. **complete**: `complete_ra` and the difference in means.
2. **complete with control for block**: `complete_ra` and OLS on $Y \sim Z + \mathrm{factor}(\mathrm{blocks})$, the same two bins as the blocked designs. That is regression adjustment on bins, not analyze-as-you-randomize. Complete assignment can empty a block; then $Z$ is aliased with the block dummy and the fit falls back to DiM.
3. **block with blocks**: `block_ra` and a blocked difference in means on those same blocks.
4. **block with blocks and linear control**: `block_ra` and `lm_robust(Y ~ Z + factor(blocks) + x, se_type = "HC2")`. Additive linear $x$ on top of the block factor.
5. **formula**: `balanced_ra(formula = ~ x)` and unadjusted DiM.
6. **formula with linear control**: the same assignment and Lin's (2013) adjustment on $x$: `lm_lin(Y ~ Z, covariates = ~ x)`. Cube-on-X flies on an intercept plus $x$. Lin uses those same columns.

Difference-in-means is still unbiased under constant $p$ for cube-on-X. It is not the paired estimator. Standard errors are HC2 (`lm_robust` / `lm_lin`).

$Y$ is $x^2$. Lin on linear $x$ is misspecified for the conditional expectation. It is still the design-paired estimator.

```{r}
set.seed(20260822)

# Complete RA can empty a block; then Z is aliased and the fit falls back to DiM.
ols_blocks <- function(data) {
  fit <- lm_robust(Y ~ Z + factor(blocks), data = data, se_type = "HC2")
  if (is.na(coef(fit)[["Z"]])) {
    fit <- lm_robust(Y ~ Z, data = data, se_type = "HC2")
  }
  tidy(fit)[tidy(fit)$term == "Z", ]
}

pop <-
  declare_model(
    N = 10,
    x = 1:N,
    blocks = cut(x, breaks = c(0, 5, 10), labels = c("1-5", "6-10"),
                 include.lowest = TRUE),
    Y_Z_0 = x^2,
    Y_Z_1 = x^2 + 10
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0))

reveal <- declare_measurement(Y = reveal_outcomes(Y ~ Z))

design_complete <- pop +
  declare_assignment(Z = complete_ra(10)) +
  reveal +
  declare_estimator(Y ~ Z, se_type = "HC2", inquiry = "ATE", label = "Simple") +
  declare_estimator(Y ~ Z + x, se_type = "HC2", inquiry = "ATE", label = "Simple + control", term = "Z")

design_blocks <- pop +
  declare_assignment(Z = block_ra(blocks = blocks)) +
  reveal +
  declare_estimator(Y ~ Z  + factor(blocks), se_type = "HC2",  term = "Z",
                    inquiry = "ATE", label = "blocked DiM") +
  declare_estimator(Y ~ Z + factor(blocks) + x, .method = lm_robust, term = "Z",
                    se_type = "HC2", inquiry = "ATE", label = "blocked + x")

design_formula <- pop +
  declare_assignment(Z = balanced_ra(formula = ~ x)) +
  reveal +
  declare_estimator(Y ~ Z, inquiry = "ATE", label = "Formula", se_type = "HC2",) +
  declare_estimator(Y ~ Z + x, term = "Z",
                    inquiry = "ATE", label = "Formula + control", se_type = "HC2")
```


```{r}
diagnose_designs(
  list(
    complete = design_complete,
    block = design_blocks,
    formula = design_formula
  ), sims = 1000
) |> reshape_diagnosis() |> dplyr::select(Estimator, Bias, Power) |>
knitr::kable(digits = 2)

```

Controlling for the two bins after complete assignment cuts RMSE from 22.5 to 14.5. Blocked assignment plus blocked DiM is near 12. Cube-on-X without Lin is near 14. Those three sit together. The extra is linear control for $x$: blocks plus $x$ is near 5.8; cube-on-X plus Lin is near 3.8. With $x$ this prognostic, most of the drop from 22.5 to 3.8 is from controlling, not from the assignment. The design pairing is the last step.

The complete-plus-block-dummies estimator is not the blocked DiM. OLS with a block factor precision-weights toward the quieter half of $x$ (the low values, where $x^2$ varies less). That is regression adjustment on bins, not analyze-as-you-randomize.

The outcome is $x^2$. Cube-on-X holds the linear total $\sum x Z$ near its target. It does not hold $\sum x^2 Z$. Lin still adjusts for linear $x$, so the CEF is misspecified. On this monotone $x$, $x$ and $x^2$ move together, and the paired estimator still has the smallest RMSE. Unadjusted DiM after cube-on-X is not a free lunch when $Y$ is nonlinear in $x$.

# 5. What to take away

Use `formula = ~ x` when the object you care about is a linear-in-$x$ total. Use `blocks` when you care about treated counts inside discrete groups. Do not pass both. Pair the estimator with the design as the default: block dummies with blocked DiM, $x$ in the formula with Lin (or OLS on those same columns of $X$). Section 4 also reports misspecified and extra controls because $x$ is so strong. With constant $p$, `complete_ra` is the count-only baseline. With unit-varying $p$, that baseline is `balanced_ra()` with no formula; `complete_ra` is a different design. Multi-arm `prob_unit_each` with `formula` is not implemented.

# References

Deville, J.-C. and Tillé, Y. (2004). Efficient balanced sampling: the cube method. *Biometrika* 91(4), 893–912. <https://doi.org/10.1093/biomet/91.4.893>

Chauvet, G. and Tillé, Y. (2006). A fast algorithm for balanced sampling. *Computational Statistics* 21(1), 53–62. <https://doi.org/10.1007/s00180-006-0250-2>

Davezies, L., Hollard, G. and Vergara Merino, P. (2024). Revisiting randomization with the cube method. <https://arxiv.org/abs/2407.13613>

Harshaw, C., Sävje, F., Spielman, D. A. and Zhang, P. (2024). Balancing covariates in randomized experiments with the Gram--Schmidt walk design. *Journal of the American Statistical Association* 119(548), 2934–2946. <https://doi.org/10.1080/01621459.2023.2285474>

Higgins, M. J., Sävje, F. and Sekhon, J. S. (2016). Improving massive experiments with threshold blocking. *Proceedings of the National Academy of Sciences* 113(27), 7369–7376. <https://doi.org/10.1073/pnas.1510504113>

Lin, W. (2013). Agnostic notes on regression adjustments to experimental data: Reexamining Freedman's critique. *Annals of Applied Statistics* 7(1), 295–318. <https://doi.org/10.1214/12-AOAS583>

Morgan, K. L. and Rubin, D. B. (2012). Rerandomization to improve covariate balance in experiments. *Annals of Statistics* 40(2), 1263–1282. <https://doi.org/10.1214/12-AOS1008>

Schneider, S. O. and Schlather, M. (2017). Rerandomization to improve covariate balance by minimizing the MSE of a treatment effect estimator. Max Planck Institute for Research on Collective Goods. <https://www.econstor.eu/handle/10419/161931>
