---
title: "How randomizr Generates Random Assignments"
author: "Alexander Coppock"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How randomizr Generates Random Assignments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
set.seed(20260101)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes the algorithms that `randomizr` uses to generate random assignments. Understanding these algorithms clarifies what the package guarantees (exact marginal probabilities, correct distributions over assignments) and why the package is fast enough for use inside DeclareDesign simulations with many thousands of draws.

## The core requirement: correct distributions

A random assignment procedure is characterized by a probability distribution over the set of all possible assignments. For an experiment with N units and two conditions, there are $2^N$ possible assignments. The procedure is correct if and only if each assignment is drawn with the right probability. For complete random assignment with m treated units, that probability is $1 / \binom{N}{m}$ for every assignment with exactly m treated units and 0 for all others. The algorithms below produce exactly this distribution.

## Simple random assignment

`simple_ra` assigns each unit independently. Each unit draws a condition from a multinomial distribution with probabilities `prob_each`. The algorithm is the simplest in the package: generate N independent uniform random numbers and threshold them to select conditions. The implementation uses `vsample`, a vectorized C routine that draws N multinomial samples in a single pass.

Simple random assignment is the right choice when units arrive in a stream and the total N is not known in advance. In all other settings, `complete_ra` is preferred because fixing the count in each condition removes a source of variability and narrows the sampling distribution of the estimator.

```{r}
library(randomizr)

# Each unit assigned independently; counts vary across draws
set.seed(1); table(simple_ra(N = 100))
set.seed(2); table(simple_ra(N = 100))
```

## Complete random assignment

`complete_ra` assigns exactly a fixed number of units to each condition. In the two-arm case with m treated units, every subset of size m is equally likely. The algorithm uses a partial Knuth shuffle (also called the Fisher-Yates shuffle): starting from a sequence of unit indices, swap index 1 with a randomly chosen index in positions 1 through N, then swap index 2 with a randomly chosen index in positions 2 through N, and so on for m steps. The first m positions after the partial shuffle identify the treated units. Drawing only m uniforms (rather than N) is both faster and sufficient.

```{r}
# Counts are fixed: exactly 50 treated, 50 control on every draw
set.seed(1); table(complete_ra(N = 100))
set.seed(2); table(complete_ra(N = 100))
```

### Stochastic rounding when prob implies a fractional count

When a researcher specifies a treatment probability rather than an exact count, the implied count `N * prob` is not always an integer. `complete_ra` resolves this with stochastic rounding: it assigns either `floor(N * prob)` or `ceiling(N * prob)` units to treatment, drawing between the two values with probability chosen so that the marginal probability of assignment is exactly `prob`. Specifically, if the fractional part of `N * prob` is `f`, then `ceiling(N * prob)` units are treated with probability `f` and `floor(N * prob)` units are treated with probability `1 - f`.

```{r}
# prob = 0.333, N = 100: 33 or 34 treated, averaging to exactly 0.333
counts <- replicate(10000, sum(complete_ra(N = 100, prob = 1/3) == 1))
mean(counts)   # very close to 33.33
```

Multi-arm designs use the same principle: first assign `floor(N * prob_each)` units to each condition, then use a single round of simple random assignment among the remaining units, with probabilities calibrated so that the overall probability of assignment equals `prob_each` exactly.

## Block random assignment

`block_ra` applies complete random assignment separately within each block. The statistical motivation is precision: if the blocking variable is correlated with potential outcomes, then guaranteeing that each block contributes both treated and control observations rules out the unlucky assignments that would otherwise generate estimates far from the true average treatment effect. The precision gain scales with the correlation between the block variable and potential outcomes.

### The two-arm fast path: counting sort plus per-block Knuth shuffle in C++

For two-arm designs, `block_ra` calls `block_assign_cpp`, a C++ routine compiled via Rcpp. The algorithm has two phases.

**Phase 1: counting sort.** Group unit indices into contiguous runs by block. First count the number of units in each block (one pass over N). Compute prefix sums to find where each block's run starts in the output array. Then fill in the output array: one more pass over N places each unit's index into the correct position. Total work: O(N) with no comparisons and no memory allocation beyond the output array.

**Phase 2: partial Knuth shuffle within each block.** For block $b$ with $n_b$ units and $m_b$ to treat, shuffle the first $m_b$ positions of that block's run in the output array using the partial Fisher-Yates shuffle described above. Mark those positions as treated. Total RNG draws across all blocks: $\sum_b m_b \leq N/2$ on average for balanced designs.

The total cost is O(N) for the sort plus O(sum of m_b) for the shuffles. The counting sort is several times faster than the alternatives: calling R's `order()` on a sort key, which uses a comparison sort at O(N log N), or calling a separate `complete_ra` for each block, which pays per-block R dispatch overhead.

```{r}
blocks <- rep(c("A", "B", "C"), times = c(50, 100, 200))

# Exactly half of each block treated by default
Z <- block_ra(blocks = blocks)
table(blocks, Z)
```

### Declaration caching for simulation

When a random assignment is declared with `declare_ra`, the integer encoding of the block variable (`as.integer(as.factor(blocks))`) is precomputed once and stored in the declaration object. `conduct_ra` passes this cached encoding directly to `block_ra` as an internal argument, avoiding the `as.factor` call on every simulation draw. For a 10,000-iteration simulation with 200 blocks this saves roughly 0.2 seconds.

```{r}
decl <- declare_ra(blocks = blocks)

# conduct_ra reuses the precomputed block encoding on every draw
Z1 <- conduct_ra(decl)
Z2 <- conduct_ra(decl)
all(table(blocks, Z1) == table(blocks, Z2))  # same counts, different assignments
```

## Cluster random assignment

`cluster_ra` assigns entire clusters to conditions. All units in a cluster share the same treatment status. The algorithm reduces the problem to complete (or simple) random assignment at the cluster level: extract the unique clusters, apply `complete_ra` to those clusters, then expand the cluster-level assignments back to the unit level.

Clustering is appropriate when treatment is an inherently group-level intervention (a classroom program, a village-level policy) or when interference within groups makes individual-level assignment infeasible. The precision cost is real: the effective sample size is the number of clusters rather than the number of units, so clustering typically increases the sampling variability of treatment effect estimates. The precision loss grows with the intra-cluster correlation in potential outcomes.

```{r}
clusters <- rep(letters[1:10], each = 20)  # 10 clusters of 20 units each

Z <- cluster_ra(clusters = clusters)
# All units in a cluster share the same assignment
tapply(Z, clusters, unique)
```

## Block and cluster random assignment

`block_and_cluster_ra` combines blocking and clustering. Clusters are grouped into blocks, and complete random assignment is applied to clusters within each block. This design captures the precision benefits of blocking (ruling out unlucky cross-block imbalances) while respecting the constraint that treatment must be assigned at the cluster level.

```{r}
# 4 blocks, each containing 5 clusters of 10 units
clusters <- rep(1:20, each = 10)
blocks <- rep(c("W", "X", "Y", "Z"), each = 5)[clusters]

Z <- block_and_cluster_ra(clusters = clusters, blocks = blocks)
# Cluster assignments within each block
with(data.frame(Z, blocks, clusters),
     tapply(Z, list(blocks, clusters), unique))
```

## Balanced random assignment

`balanced_ra` assigns units when the probability of assignment varies from unit to unit, and it holds the number assigned to each condition at the floor or the ceiling of that condition's expected count. Neither of the algorithms above will do that job. Complete random assignment fixes the counts but requires one probability shared by every unit; simple random assignment allows the probabilities to vary but fixes no counts at all. `balanced_ra` is the design that does both, and it is cube/pivotal assignment on the same interface as `complete_ra` and `block_ra`.

The input is an $N \times k$ matrix $\pi$ of assignment probabilities. Row $i$ sums to 1, so each unit receives exactly one condition. Column $j$ sums to a target $t_j = \sum_i \pi_{ij}$. The output is a 0-1 matrix $Z$ with the same row sums and with $E[Z] = \pi$, whose column sums are required to be as close to $t$ as integer arithmetic allows: each realized count $n_j$ satisfies $n_j \in \{\lfloor t_j \rfloor, \lceil t_j \rceil\}$.

### Cycles, then paths

Picture the algorithm as a flow on the bipartite graph of units and conditions, with an edge wherever a cell is still fractional. A cycle in that graph is a closed walk that alternates between units and conditions. Pushing a constant $\delta$ forward along the even edges and backward along the odd edges leaves every row total and every column total unchanged, and a large enough $\delta$ drives at least one cell to 0 or 1. Repeat until no cycle remains. That is the *flight* phase, in the terminology of the cube method.

What remains is a forest. Every unit still has either no fractional cells or at least two of them: a unit with a single fractional cell would need that cell to be an integer, because its other cells are already 0 or 1 and the row sums to 1. Every leaf of the forest is therefore a condition, and a maximal path runs from one condition to another. Pushing along such a path preserves every row total and every interior column total; only the two conditions at the ends move. That is the *landing* phase.

Each push is a fair bet. If the largest feasible up-step is $d^{+}$ and the largest feasible down-step is $d^{-}$, the algorithm takes the up-step with probability $d^{-}/(d^{+}+d^{-})$, which makes the current matrix a martingale:

$$E[Z^{(t+1)} \mid Z^{(t)}] = Z^{(t)}.$$

Unrolling the martingale gives $E[Z] = \pi$. The inclusion probabilities are exactly the probabilities supplied, not an approximation to them. The same degree bound that appears in dependent rounding explains the floor-or-ceiling counts: when the walk finishes, each vertex of the bipartite graph has integer degree equal to the floor or the ceiling of the fractional degree it started with. For a condition that degree is the number of units assigned to it; for a unit it is 1.

Two arms are a special case worth naming, because it is the one most designs are in. The transportation graph is then a forest of stars, one edge per unit, so every maximal path runs from one unit through its block to another unit, and the general walk collapses to the pivotal method of Deville and Tillé (1998): transfer mass between two units until one of them hits 0 or 1, with the direction chosen so that the step is a martingale and the total stays exact. Three or more arms take the windowed flight of Chauvet and Tillé (2006), whose cost is linear in $N$ because only as many units as there are constraints need to be examined at once. A window of $k$ fractional units induces at least $2k$ edges on at most $2k$ nodes, and a graph with at least as many edges as nodes is not a forest, so the window always holds a cycle until too few fractional units remain.

### What the papers guarantee, and what this implementation guarantees

The distinction matters, because the two are not the same. Deville and Tillé (2004) supply the martingale and the flight/landing split for balanced sampling in general, where the balancing constraints are linear equalities: in survey sampling, that the Horvitz–Thompson totals of auxiliary variables match known population totals. Landing may violate a constraint that was dropped to reach a vertex, and the paper does not claim that every balancing constraint survives all the way to the sample. `balanced_ra` applies that method to one particular constraint system, the transportation polytope of nonnegative $N \times k$ matrices with row sums 1 and column sums $t$. Cycle moves are the kernel directions; path moves are landing. Nothing in the papers, by itself, says that a blocked assignment problem will come out tight both within each block and overall.

What this implementation guarantees is that every unit receives exactly one condition, that each unit's probability of each condition equals the supplied $\pi_{ij}$, and that condition counts are the floor or the ceiling of the target at a stated level:

- without `blocks`, the overall counts are tight;
- with `blocks`, the within-block counts are always tight;
- with `blocks` and two arms, the overall counts are tight as well;
- with `clusters`, the tight counts are counts of clusters, not of units.

The two-arm overall-and-block result is an implementation choice rather than a named theorem that the cube always does both. After a first pivotal pass inside each block, at most one unit per block is still fractional. A second pass pairs those leftovers as a single group. Pairing preserves their total, so the overall count stays tight, and each leftover is still sent to 0 or 1, so no block moves by more than one from its target. Independent Bernoulli rounding of each leftover would have kept each block tight and let the overall count drift by up to the number of blocks.

### Caveats

`balanced_ra` is experimental and its interface may change.

The leftover pairing does not extend to three or more arms. A block can keep several leftover units, and landing those leftovers as one global group could push a within-block arm count more than one away from its target, so the implementation lands each block on its own instead. Within-block counts stay at the floor or the ceiling, and the overall count is the sum of independent block landings, which can move by more than one when several remainders fall the same way. Three blocks of two units, three arms, every probability $1/3$: each tight block gives each arm 0 or 1, the overall target is 2, and the three blocks can put an arm at 0 or at 3.

```{r}
set.seed(16)
P <- matrix(1/3, 6, 3)
bl <- rep(1:3, each = 2)
Z <- replicate(2000, as.integer(as.character(
  balanced_ra(prob_unit_each = P, blocks = bl, conditions = 1:3))))
table(colSums(Z == 1))
```

Arm 1's overall count is not confined to $\{1, 2\}$. The coupling that would confine it is feasible, since Hamilton's method of apportionment would give each arm to exactly two of the three blocks, and this implementation does not do it when $k > 2$.

There is no residual condition if the rows of `prob_unit_each` do not sum to 1, and no interpretation of a probability greater than 1 as an expected count. Both are refused. `check_inputs = FALSE` skips those checks, and it is meant for simulation loops whose probabilities have already been verified rather than as a way to pass illegal designs.

Holding assignment *counts* against their targets is not the same thing as balancing a continuous covariate. Passing `formula` runs cube-on-X on a model matrix, and `blocks` tightens counts within discrete groups; the two are different constraints, and the function refuses both at once. See `vignette("balanced_ra_covariates", package = "randomizr")`.

The practical summary is short. Use `balanced_ra` when probabilities vary across units and the assigned counts should stay as close to their targets as arithmetic allows. Trust overall tightness without `blocks`, and both overall and within-block tightness with two arms. Do not trust overall tightness with three or more arms and `blocks`. The full walk-through, with one two-arm draw and one four-by-three draw shown cell by cell, is in `vignette("balanced_ra", package = "randomizr")`. HC2 coverage under these designs is in the article "HC2 coverage under balanced assignment", on the package site at <https://declaredesign.org/r/randomizr/articles/>.

## Probability calculations

For every design type, a companion `*_probabilities` function returns the matrix of assignment probabilities, one row per unit and one column per arm. Inverse-probability-weighted estimators use these probabilities, and `declare_ra` stores them in the declaration object.

For complete random assignment, the probability of assignment to treatment is simply `m / N`. For block random assignment, it is `m_b / n_b` within each block $b$, where `m_b` is the number treated and `n_b` is the block size. For cluster designs, every unit in a cluster inherits the cluster's assignment probability. `declare_ra` precomputes these probabilities once at declaration time, so `obtain_condition_probabilities` is fast regardless of design complexity.

```{r}
# Assignment probabilities for a blocked design with varying block probabilities
prob_mat <- block_ra_probabilities(
  blocks = blocks,
  block_prob = c(0.2, 0.4, 0.6, 0.8)
)
head(prob_mat)
```

## References

Chauvet, G. and Tillé, Y. (2006). A fast algorithm for balanced sampling. *Computational Statistics* 21(1), 53–62. <https://doi.org/10.1007/s00180-006-0250-2>

Deville, J.-C. and Tillé, Y. (1998). Unequal probability sampling without replacement through a splitting method. *Biometrika* 85(1), 89–101. <https://doi.org/10.1093/biomet/85.1.89>

Deville, J.-C. and Tillé, Y. (2004). Efficient balanced sampling: the cube method. *Biometrika* 91(4), 893–912. <https://doi.org/10.1093/biomet/91.4.893>

Deville, J.-C. and Tillé, Y. (2005). Variance approximation under balanced sampling. *Journal of Statistical Planning and Inference* 128(2), 569–591. <https://doi.org/10.1016/j.jspi.2003.11.011>. `balanced_ra` does not implement these estimators.
