---
title: "How randomizr Generates Random Assignments"
author: "Alexander Coppock"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How randomizr Generates Random Assignments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
set.seed(20260101)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes the algorithms that `randomizr` uses to generate random assignments. Understanding these algorithms clarifies what the package guarantees (exact marginal probabilities, correct distributions over assignments) and why the package is fast enough for use inside DeclareDesign simulations with many thousands of draws.

## The core requirement: correct distributions

A random assignment procedure is characterized by a probability distribution over the set of all possible assignments. For an experiment with N units and two conditions, there are $2^N$ possible assignments. The procedure is correct if and only if each assignment is drawn with the right probability. For complete random assignment with m treated units, that probability is $1 / \binom{N}{m}$ for every assignment with exactly m treated units and 0 for all others. The algorithms below produce exactly this distribution.

## Simple random assignment

`simple_ra` assigns each unit independently. Each unit draws a condition from a multinomial distribution with probabilities `prob_each`. The algorithm is the simplest in the package: generate N independent uniform random numbers and threshold them to select conditions. The implementation uses `vsample`, a vectorized C routine that draws N multinomial samples in a single pass.

Simple random assignment is the right choice when units arrive in a stream and the total N is not known in advance. In all other settings, `complete_ra` is preferred because fixing the count in each condition removes a source of variability and narrows the sampling distribution of the estimator.

```{r}
library(randomizr)

# Each unit assigned independently; counts vary across draws
set.seed(1); table(simple_ra(N = 100))
set.seed(2); table(simple_ra(N = 100))
```

## Complete random assignment

`complete_ra` assigns exactly a fixed number of units to each condition. In the two-arm case with m treated units, every subset of size m is equally likely. The algorithm uses a partial Knuth shuffle (also called the Fisher-Yates shuffle): starting from a sequence of unit indices, swap index 1 with a randomly chosen index in positions 1 through N, then swap index 2 with a randomly chosen index in positions 2 through N, and so on for m steps. The first m positions after the partial shuffle identify the treated units. Drawing only m uniforms (rather than N) is both faster and sufficient.

```{r}
# Counts are fixed: exactly 50 treated, 50 control on every draw
set.seed(1); table(complete_ra(N = 100))
set.seed(2); table(complete_ra(N = 100))
```

### Stochastic rounding when prob implies a fractional count

When a researcher specifies a treatment probability rather than an exact count, the implied count `N * prob` is not always an integer. `complete_ra` resolves this with stochastic rounding: it assigns either `floor(N * prob)` or `ceiling(N * prob)` units to treatment, drawing between the two values with probability chosen so that the marginal probability of assignment is exactly `prob`. Specifically, if the fractional part of `N * prob` is `f`, then `ceiling(N * prob)` units are treated with probability `f` and `floor(N * prob)` units are treated with probability `1 - f`.

```{r}
# prob = 0.333, N = 100: 33 or 34 treated, averaging to exactly 0.333
counts <- replicate(10000, sum(complete_ra(N = 100, prob = 1/3) == 1))
mean(counts)   # very close to 33.33
```

Multi-arm designs use the same principle: first assign `floor(N * prob_each)` units to each condition, then use a single round of simple random assignment among the remaining units, with probabilities calibrated so that the overall probability of assignment equals `prob_each` exactly.

## Block random assignment

`block_ra` applies complete random assignment separately within each block. The statistical motivation is precision: if the blocking variable is correlated with potential outcomes, then guaranteeing that each block contributes both treated and control observations rules out the unlucky assignments that would otherwise generate estimates far from the true average treatment effect. The precision gain scales with the correlation between the block variable and potential outcomes.

### The two-arm fast path: counting sort plus per-block Knuth shuffle in C++

For two-arm designs, `block_ra` calls `block_assign_cpp`, a C++ routine compiled via Rcpp. The algorithm has two phases.

**Phase 1: counting sort.** Group unit indices into contiguous runs by block. First count the number of units in each block (one pass over N). Compute prefix sums to find where each block's run starts in the output array. Then fill in the output array: one more pass over N places each unit's index into the correct position. Total work: O(N) with no comparisons and no memory allocation beyond the output array.

**Phase 2: partial Knuth shuffle within each block.** For block $b$ with $n_b$ units and $m_b$ to treat, shuffle the first $m_b$ positions of that block's run in the output array using the partial Fisher-Yates shuffle described above. Mark those positions as treated. Total RNG draws across all blocks: $\sum_b m_b \leq N/2$ on average for balanced designs.

The total cost is O(N) for the sort plus O(sum of m_b) for the shuffles. The counting sort is several times faster than the alternatives: calling R's `order()` on a sort key, which uses a comparison sort at O(N log N), or calling a separate `complete_ra` for each block, which pays per-block R dispatch overhead.

```{r}
blocks <- rep(c("A", "B", "C"), times = c(50, 100, 200))

# Exactly half of each block treated by default
Z <- block_ra(blocks = blocks)
table(blocks, Z)
```

### Declaration caching for simulation

When a random assignment is declared with `declare_ra`, the integer encoding of the block variable (`as.integer(as.factor(blocks))`) is precomputed once and stored in the declaration object. `conduct_ra` passes this cached encoding directly to `block_ra` as an internal argument, avoiding the `as.factor` call on every simulation draw. For a 10,000-iteration simulation with 200 blocks this saves roughly 0.2 seconds.

```{r}
decl <- declare_ra(blocks = blocks)

# conduct_ra reuses the precomputed block encoding on every draw
Z1 <- conduct_ra(decl)
Z2 <- conduct_ra(decl)
all(table(blocks, Z1) == table(blocks, Z2))  # same counts, different assignments
```

## Cluster random assignment

`cluster_ra` assigns entire clusters to conditions. All units in a cluster share the same treatment status. The algorithm reduces the problem to complete (or simple) random assignment at the cluster level: extract the unique clusters, apply `complete_ra` to those clusters, then expand the cluster-level assignments back to the unit level.

Clustering is appropriate when treatment is an inherently group-level intervention (a classroom program, a village-level policy) or when interference within groups makes individual-level assignment infeasible. The precision cost is real: the effective sample size is the number of clusters rather than the number of units, so clustering typically increases the sampling variability of treatment effect estimates. The precision loss grows with the intra-cluster correlation in potential outcomes.

```{r}
clusters <- rep(letters[1:10], each = 20)  # 10 clusters of 20 units each

Z <- cluster_ra(clusters = clusters)
# All units in a cluster share the same assignment
tapply(Z, clusters, unique)
```

## Block and cluster random assignment

`block_and_cluster_ra` combines blocking and clustering. Clusters are grouped into blocks, and complete random assignment is applied to clusters within each block. This design captures the precision benefits of blocking (ruling out unlucky cross-block imbalances) while respecting the constraint that treatment must be assigned at the cluster level.

```{r}
# 4 blocks, each containing 5 clusters of 10 units
clusters <- rep(1:20, each = 10)
blocks <- rep(c("W", "X", "Y", "Z"), each = 5)[clusters]

Z <- block_and_cluster_ra(clusters = clusters, blocks = blocks)
# Cluster assignments within each block
with(data.frame(Z, blocks, clusters),
     tapply(Z, list(blocks, clusters), unique))
```

## Probability calculations

For every design type, a companion `*_probabilities` function returns the matrix of assignment probabilities, one row per unit and one column per arm. Inverse-probability-weighted estimators use these probabilities, and `declare_ra` stores them in the declaration object.

For complete random assignment, the probability of assignment to treatment is simply `m / N`. For block random assignment, it is `m_b / n_b` within each block $b$, where `m_b` is the number treated and `n_b` is the block size. For cluster designs, every unit in a cluster inherits the cluster's assignment probability. `declare_ra` precomputes these probabilities once at declaration time, so `obtain_condition_probabilities` is fast regardless of design complexity.

```{r}
# Assignment probabilities for a blocked design with varying block probabilities
prob_mat <- block_ra_probabilities(
  blocks = blocks,
  block_prob = c(0.2, 0.4, 0.6, 0.8)
)
head(prob_mat)
```
