---
title: "Speed of balanced_ra relative to complete and blocked assignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speed of balanced_ra relative to complete and blocked assignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
set.seed(20260822)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.2
)
```

```{r}
library(randomizr)
library(ggplot2)
```

`balanced_ra()` is the general assignment function when probabilities may vary by unit. When every unit has the same probability, `complete_ra()`, `block_ra()`, and `block_and_cluster_ra()` already implement the same designs. This note times the general function against those specialized functions on the equal-probability cases.

Three results from the grid below. First, on two-arm complete assignment `complete_ra()` is faster once $N$ is a few thousand; at $N = 200$ the two are close and the ranking can flip from run to run. Second, blocked assignment is close to even: `block_ra()` uses a specialized Knuth-shuffle path, and the cube walk in `balanced_ra()` does not pull away. Third, on blocked-and-clustered assignment `balanced_ra()` is the faster of the two once validation is switched off on both sides, by a factor of roughly three to four, because `block_and_cluster_ra()` rebuilds the cluster-to-block map on every draw. At the settings a user actually gets, with validation on, that result reverses: see [Reading the ratios](#reading-the-ratios).

`balanced_ra()` is the general tool. The specialized functions exist to be faster on the equal-$p$ cases. Use them when they apply.

A later section times two-arm cube-on-X against `BalancedSampling::cube` and `sampling::samplecube`. `BalancedSampling` is faster on that problem. That is not a reason to drop the C++ for the count path, leftover pairing, or multi-arm assignment.

# What is timed

Each cell draws the same two-arm design twice, once with the specialized function and once with `balanced_ra()`, and records elapsed seconds for a fixed number of draws. The ratio is balanced time divided by specialized time. A ratio above 1 means `balanced_ra()` is slower.

The timed loop sets `check_inputs = FALSE` on **both** sides, so the comparison is assignment cost rather than validation cost. Symmetry here is not cosmetic. Every version of this note before randomizr 2.0.1 switched validation off on `balanced_ra()` only, because `check_inputs = FALSE` did not work at all for `block_ra()`, `cluster_ra()`, or `block_and_cluster_ra()`: the same code path that validated the arguments also derived `num_arms` and `conditions`, so skipping the validation skipped the derivation and the call failed. The two sides could not be timed on equal terms until that was fixed, and the conclusion it produced for blocked-and-clustered designs was the wrong way round. Probability is the default $0.5$ everywhere. Blocks are equal in size. Cluster designs nest clusters of size 10 inside those blocks.

```{r}
Ns <- c(200L, 2000L, 10000L)
n_blocks_grid <- c(1L, 10L, 50L, 200L)
cluster_size <- 10L

# Enough draws that elapsed times sit above a typical Windows clock tick.
# A longer run with more draws produced the same ranking.
nrep_for <- function(N) {
  if (N <= 200L) 800L else if (N <= 2000L) 250L else 80L
}

time_n <- function(n, f) {
  system.time(for (i in seq_len(n)) f())[["elapsed"]]
}

row_out <- function(family, N, n_blocks, cluster_size, nrep, specialized, balanced) {
  data.frame(
    family = family,
    N = N,
    n_blocks = n_blocks,
    cluster_size = cluster_size,
    nrep = nrep,
    specialized = specialized,
    balanced = balanced,
    ratio = balanced / specialized,
    stringsAsFactors = FALSE
  )
}

# Warm-up so compilation and first-call costs sit outside the timed loops.
invisible(complete_ra(20))
invisible(block_ra(blocks = rep(1:2, each = 10)))
invisible(block_and_cluster_ra(blocks = rep(1:2, each = 20),
                              clusters = rep(1:8, each = 5)))
invisible(balanced_ra(20, check_inputs = FALSE))

rows <- list()

for (N in Ns) {
  n <- nrep_for(N)
  rows[[length(rows) + 1L]] <- row_out(
    "complete", N, NA_integer_, NA_integer_, n,
    time_n(n, function() complete_ra(N, check_inputs = FALSE)),
    time_n(n, function() balanced_ra(N, check_inputs = FALSE))
  )
}

for (N in Ns) {
  for (nb in n_blocks_grid) {
    if (N %% nb != 0L) next
    blocks <- rep(seq_len(nb), each = N / nb)
    n <- nrep_for(N)
    rows[[length(rows) + 1L]] <- row_out(
      "blocked", N, nb, NA_integer_, n,
      time_n(n, function() block_ra(blocks = blocks, check_inputs = FALSE)),
      time_n(n, function() balanced_ra(blocks = blocks, check_inputs = FALSE))
    )
  }
}

for (N in Ns) {
  if (N %% cluster_size != 0L) next
  n_cl <- N / cluster_size
  clusters <- rep(seq_len(n_cl), each = cluster_size)
  for (nb in n_blocks_grid) {
    if (n_cl %% nb != 0L) next
    blocks <- rep(seq_len(nb), each = N / nb)
    n <- nrep_for(N)
    rows[[length(rows) + 1L]] <- row_out(
      "block_cluster", N, nb, cluster_size, n,
      time_n(n, function() block_and_cluster_ra(blocks = blocks, clusters = clusters,
                                               check_inputs = FALSE)),
      time_n(n, function() balanced_ra(blocks = blocks, clusters = clusters,
                                      check_inputs = FALSE))
    )
  }
}

times <- do.call(rbind, rows)
times$ms_specialized <- 1000 * times$specialized / times$nrep
times$ms_balanced <- 1000 * times$balanced / times$nrep
```

The complete grid is $N \in \{200, 2000, 10000\}$. The blocked grid crosses those $N$ with $1$, $10$, $50$, and $200$ blocks, all of which divide $N$. The cluster grid uses the same $N$ and block counts whenever the number of clusters is divisible by the number of blocks, so $N = 200$ drops the $50$- and $200$-block cells.

# Times

Elapsed seconds are totals over `nrep` draws. The ratio is balanced / specialized. Absolute times vary by machine; the ranking is the object of interest.

```{r}
show_times <- function(d) {
  out <- d[, c("N", "n_blocks", "nrep", "specialized", "balanced", "ratio")]
  out$specialized <- round(out$specialized, 3)
  out$balanced <- round(out$balanced, 3)
  out$ratio <- round(out$ratio, 2)
  rownames(out) <- NULL
  out
}

show_times(times[times$family == "complete", ])
show_times(times[times$family == "blocked", ])
show_times(times[times$family == "block_cluster", ])
```

Figure 1 plots milliseconds per draw against $N$. Each panel is one design family. For blocked and clustered designs the series average over the block counts in the table.

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Milliseconds per draw against N. Each panel is one design family. Blocked and clustered series average over the block counts in the table. A ratio above 1 in the table means balanced_ra is slower."}
agg <- aggregate(
  cbind(ms_specialized, ms_balanced) ~ family + N,
  data = times,
  FUN = mean
)
plot_df <- rbind(
  data.frame(family = agg$family, N = agg$N,
             method = "specialized", ms = agg$ms_specialized),
  data.frame(family = agg$family, N = agg$N,
             method = "balanced_ra", ms = agg$ms_balanced)
)
plot_df$family <- factor(
  plot_df$family,
  levels = c("complete", "blocked", "block_cluster"),
  labels = c("complete_ra", "block_ra", "block_and_cluster_ra")
)
ggplot(plot_df, aes(N, ms, color = method, shape = method)) +
  geom_line() +
  geom_point() +
  scale_x_log10(breaks = Ns) +
  facet_wrap(~ family, scales = "free_y") +
  labs(x = "N", y = "ms per draw", color = NULL, shape = NULL) +
  theme_bw(base_size = 9) +
  theme(legend.position = "bottom")
```

# Reading the ratios

On complete assignment the specialized function is the faster path at large $N$. `complete_ra()` shuffles a length-$N$ vector. `balanced_ra()` walks the cube on an $N \times 2$ probability matrix. That walk is linear in $N$, but the constant is larger than a shuffle of $N$ labels. At $N = 200$ the difference is small relative to function-call noise.

On blocked assignment the two stay near a ratio of 1. `block_ra()` already has a C++ Knuth-shuffle fast path for the two-arm equal-$p$ case. Replacing it with the cube does not buy time, and usually does not cost much either.

On blocked-and-clustered assignment `balanced_ra()` is faster, by a factor of roughly three to four across this grid. `block_and_cluster_ra()` calls `tapply()` on every draw to collapse units to clusters, and `order(unlist(split()))` to put them back; on a $N = 10{,}000$ draw that setup is most of the elapsed time. `balanced_ra()` assigns at the cluster level from a collapsed probability matrix. The gap is setup, not the cube.

That result holds only because both sides have validation switched off. At default settings it reverses, and `balanced_ra()` is the slower of the two by a factor of about 1.1 at $N = 200$ rising to about 3.1 at $N = 10{,}000$. The reason is symmetric to the one above: `balanced_ra_matrix()` validates clusters with a per-draw `tapply()` over every column of the probability matrix, which is the same kind of per-draw setup cost, paid on the other side. Neither function's validation is free, and a comparison that switches one off and leaves the other on measures the switch rather than the algorithm. Cheapening that cluster validation is an open item; until it lands, a simulation loop that draws the same blocked-and-clustered design many times should set `check_inputs = FALSE` on whichever function it uses, having checked the design once outside the loop.

None of these comparisons is a reason to call `balanced_ra()` when probabilities are constant. They say only what the general function costs on the cases the specialized functions already cover. When probabilities vary by unit, those specialized functions do not apply.

The takeaway is the division of labor. Use `complete_ra()`, `block_ra()`, or `block_and_cluster_ra()` when every unit (or cluster) shares a probability. Use `balanced_ra()` when they do not.

# Cube-on-X versus the survey packages

Two-arm cube-on-X is the survey problem: the sample is the treated group. `BalancedSampling::cube` and `sampling::samplecube` already implement it. Wrapping either would avoid maintaining `src/cube_on_x.cpp` for that case.

The rest of `balanced_ra` is not that problem. The count path when `formula` is `NULL`, leftover pairing inside blocks, multi-arm assignment, and `declare_ra` have no counterpart in those packages. The APIs also differ: `pik` and `X` versus `conditions` and `prob_unit_each`. Landing rules differ. A wrapper would not control two-arm block leftover pairing. Adding either package as an Import has CRAN weight; `sampling` is the slower of the two.

The package keeps its own C++. A later option is to dispatch two-arm `formula` to `BalancedSampling::cube` when that package is installed. That is not implemented here.

The timings below are the reason the survey packages belong in the comparison, not a reason to wrap them as the implementation.

## What is timed

Each method draws a two-arm assignment with intercept plus a continuous $x$ and $p_i = 0.5$. `balanced_ra(formula = ~ x)` builds $X$ as `model.matrix(~ x)`. `BalancedSampling::cube` is called as `cube(pik, cbind(pik, x))`, the documented way to hold the sample size. `sampling::samplecube` is called on `cbind(1, x)` with `comment = FALSE` and `method = 2` (drop variables at landing, not the linear-programming landing). When $p$ is constant those two $X$ matrices span the same space as the formula. `fastflightcube` is the flight phase only; it is not a 0-1 assignment.

The timed loop sets `check_inputs = FALSE` on `balanced_ra`. $N \in \{200, 2000\}$. Median milliseconds per draw, over batches. Chunks for `BalancedSampling` and `sampling` run only if the package is installed.

```{r}
has_bs <- requireNamespace("BalancedSampling", quietly = TRUE)
has_samp <- requireNamespace("sampling", quietly = TRUE)

median_ms <- function(n_batch, nrep, f) {
  elapsed <- replicate(n_batch, {
    system.time(for (i in seq_len(nrep)) f())[["elapsed"]]
  })
  1000 * median(elapsed) / nrep
}

nrep_cube <- function(N) if (N <= 200L) 400L else 80L
Ns_cube <- c(200L, 2000L)
n_batch_cube <- 5L

cube_rows <- list()
for (N in Ns_cube) {
  x <- rnorm(N)
  pik <- rep(0.5, N)
  dat <- data.frame(x = x)
  X_int <- cbind(1, x)
  X_pik <- cbind(pik, x)
  nrep <- nrep_cube(N)

  cube_rows[[length(cube_rows) + 1L]] <- data.frame(
    N = N, method = "balanced_ra(formula = ~ x)",
    ms = median_ms(n_batch_cube, nrep, function() {
      balanced_ra(N = N, formula = ~ x, check_inputs = FALSE)
    }),
    stringsAsFactors = FALSE
  )

  if (has_bs) {
    cube_rows[[length(cube_rows) + 1L]] <- data.frame(
      N = N, method = "BalancedSampling::cube",
      ms = median_ms(n_batch_cube, nrep, function() {
        s <- BalancedSampling::cube(pik, X_pik)
        z <- integer(N)
        z[s] <- 1L
        z
      }),
      stringsAsFactors = FALSE
    )
  }

  if (has_samp) {
    nrep_s <- if (N <= 200L) 20L else 3L
    cube_rows[[length(cube_rows) + 1L]] <- data.frame(
      N = N, method = "sampling::samplecube",
      ms = median_ms(3L, nrep_s, function() {
        sampling::samplecube(X_int, pik, order = 1, comment = FALSE, method = 2)
      }),
      stringsAsFactors = FALSE
    )
    cube_rows[[length(cube_rows) + 1L]] <- data.frame(
      N = N, method = "sampling::fastflightcube",
      ms = median_ms(3L, nrep_s, function() {
        sampling::fastflightcube(X_int, pik, order = 1, comment = FALSE)
      }),
      stringsAsFactors = FALSE
    )
  }
}

cube_times <- do.call(rbind, cube_rows)
cube_times$ms <- round(cube_times$ms, 3)
```

```{r, eval = !has_bs, echo = FALSE, results = "asis"}
cat("*BalancedSampling is not installed; that series is omitted.*\n")
```

```{r, eval = !has_samp, echo = FALSE, results = "asis"}
cat("*sampling is not installed; that series is omitted.*\n")
```

```{r}
knitr::kable(
  cube_times,
  caption = "Median milliseconds per draw. Two-arm cube-on-X, intercept plus continuous $x$, $p = 0.5$. `fastflightcube` is flight only."
)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Median milliseconds per draw against N for two-arm cube-on-X. Absolute times vary by machine; the ranking is the object of interest."}
ggplot(cube_times, aes(N, ms, color = method, shape = method)) +
  geom_line() +
  geom_point() +
  scale_x_log10(breaks = Ns_cube) +
  scale_y_log10() +
  labs(x = "N", y = "ms per draw (median)", color = NULL, shape = NULL) +
  theme_bw(base_size = 9) +
  theme(legend.position = "bottom")
```

On this grid `BalancedSampling::cube` has the smallest median time. `sampling::samplecube` has the largest. `balanced_ra(formula = ~ x)` sits between them. `fastflightcube` is close to `samplecube`: the R flight, not landing, is what costs. Absolute times vary by machine; the ranking is the finding.

That ranking is a reason to consider a later dispatch of two-arm `formula` to `BalancedSampling` if it is installed. It is not a reason to replace the C++ as the implementation of `balanced_ra`. The count path, leftover pairing, and multi-arm assignment still need code this package owns.
